---
title: "Improving Data Visualizations"
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    number-sections: true
    embed-resources: true
---

# Import data

```{r } 
#| message: false
#| code-fold: TRUE

library(openintro)
library(tidyverse)
babies <- babies %>% 
  mutate(smoke = as.logical(smoke),
         parity = as.logical(parity))

xaringanExtra::use_panelset()
theme_set(theme_gray(base_size = 18))

```
 


# Scatter plot

Task:

- Using the `babies` data,   
- Map `gestation` to x-axis, `bwt` to y-axis, `smoke` to shape and color.
- Add a layer of points and set the size of the points to 4.


```{r } 
#| message: false
#| code-fold: TRUE

ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point(size = 4) 
```


# Add labels to axis using `labs()`

Task: 

- Using the `babies` data,   
- Map `gestation` to x-axis, `bwt` to y-axis, `smoke` to shape.
- Add a layer of points and set the size of the points to 4.
- [Add labels to x-axis (Gestation), y-axis (Birth Weight), and the title of the plot (Palmer babies).]{.highlight-text}

```{r } 
#| message: false
#| code-fold: TRUE

ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point(size = 4) +
  labs(x = "Gestation (days)", #<<
       y = "Birth Weight (ounces)", #<<
       title = "Baby Gestation and Birth Weight by Mother's Smoker Status") #<<
```

 
```{r } 
#| message: false
#| code-fold: TRUE
#|
ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point(size = 4) +
  labs(x = "Gestation (days)", #<<
       y = "Birth Weight (ounces)", #<<
       title = "Gestation and Birth Weight by Mother's Smoker Status") #<<
```
  
```{r } 
#| message: false
#| code-fold: TRUE
ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point() +
labs(x = "Gestation (days)",
       y = "Birth Weight (ounces)", 
       title = "Gestation and Birth Weight by Mother's Smoker Status") 
  theme_bw() #<<
  
```
  
# Themes
- `theme_gray()`

```{r } 
#| message: false
#| code-fold: TRUE
#| output-fold: TRUE
ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point() +
  labs(x = "Gestation (days)",
       y = "Birth Weight (ounces)", 
       title = "Baby Gestation and Birth Weight by Mother's Smoker Status") +
  theme_gray()
```

-`theme_bw()`

```{r } 
#| message: false
#| code-fold: TRUE
#| output-fold: TRUE
ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point() +
  labs(x = "Gestation (days)",
       y = "Birth Weight (ounces)", 
       title = "Baby Gestation and Birth Weight by Mother's Smoker Status") +
  theme_bw()
```

-`theme_classic()`

```{r } 
#| message: false
#| code-fold: TRUE
#| output-fold: TRUE
ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point() +
  labs(x = "Gestation (days)",
       y = "Birth Weight (ounces)", 
       title = "Baby Gestation and Birth Weight by Mother's Smoker Status") +
  theme_classic()
```

-`theme_dark()`

```{r } 
#| message: false
#| code-fold: TRUE
#| output-fold: TRUE
ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point() +
  labs(x = "Gestation (days)",
       y = "Birth Weight (ounces)", 
       title = "Baby Gestation and Birth Weight by Mother's Smoker Status") +
  theme_dark()
```

-`theme_minimal()`

```{r } 
#| message: false
#| code-fold: TRUE
#| output-fold: TRUE
ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point() +
  labs(x = "Gestation (days)",
       y = "Birth Weight (ounces)", 
       title = "Baby Gestation and Birth Weight by Mother's Smoker Status") +
  theme_minimal()
```

-`theme_void()`

```{r } 
#| message: false
#| code-fold: TRUE
#| output-fold: TRUE
ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point() +
  labs(x = "Gestation (days)",
       y = "Birth Weight (ounces)", 
       title = "Baby Gestation and Birth Weight by Mother's Smoker Status") +
  theme_void()
```
 

## Explore `theme()`


```{r}
?theme
```

There are more aspects of a theme that we can control. 


```{r } 
#| message: false
#| code-fold: TRUE
#| output-fold: TRUE
ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point() +
  labs(x = "Gestation (days)",
       y = "Birth Weight (ounces)", 
       title = "Baby Gestation and Birth Weight by Mother's Smoker Status") +
  theme_bw() +
  theme(title = 
          element_text(size = 12),
        axis.title = 
          element_text(size = 10,
                       face="bold"))
  
```
 
 
# Layout panels in a grid 

```{r } 
#| message: false
#| code-fold: TRUE
#| output-fold: TRUE
#| 
ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point(size = 4) +
  facet_grid(.~parity) #<<
```

 
```{r } 
#| message: false
#| code-fold: TRUE
#| output-fold: TRUE
#| 
ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point(size = 4) +
  facet_grid(parity~.) #<<
```
 
# Summary

You can do much more in ggplot. 

- Use images 
- Make maps
- Pick colors that you want
 

Check out [the ggplot flipbook](https://evamaerey.github.io/ggplot_flipbook/ggplot_flipbook_xaringan.html#1) for some inspiration. 


Additional resources:
 
- [BBC](https://bbc.github.io/rcookbook/)   
- [fivethirtyeight](https://blog.revolutionanalytics.com/2016/07/data-journalism-with-r-at-538.html?utm_content=bufferc03e0&utm_medium=social&utm_source=twitter.com&utm_campaign=buffer)  
- [NY Times](https://livefreeordichotomize.com/2021/04/07/nytimes-map-how-to/)
