---
title: "Introduction to data wrangling"
author: "Dr. Mine Dogucu"
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    number-sections: true
    embed-resources: true
---

# Import Data
 
```{r}
#| message: false
#| code-fold: TRUE

library(tidyverse)
options(scipen = 999)

arthritis <- read_csv("https://raw.githubusercontent.com/cosmos-uci-dshs/data/main/RheumArth_Tx_AgeComparisons.csv") |>
  janitor::clean_names() |>
  mutate(sex = case_when(sex == 0 ~ "female",
                         sex == 1 ~ "male"),
         sex = as.factor(sex),
         age_gp = case_when(age_gp == 1 ~ "control",
                            age_gp == 2 ~ "elderly"),
         age_gp = as.factor(age_gp),
         cdai_yn = case_when(cdai_yn == 1 ~ "no",
                             cdai_yn == 2 ~ "yes"),
         cdai_yn = as.factor(cdai_yn)) |>
  select(age, age_gp, sex, yrs_from_dx, cdai)
```
 

## Data Overview

We are working with the `arthritis` dataset. Here is a quick look at its structure:

```{r}
glimpse(arthritis)
```
 

# Mutating existing variables

We use `mutate()` to create or modify variables in a data frame.


## Creating a numeric variable: `age_months`

**Goal:** Create a new variable called `age_months` that represents `age` in months.

```{r}
#| code-fold: TRUE
arthritis |>
  mutate(age_months = age * 12)
```

Note that this does not change the original `arthritis` object. To save the result, assign it:

```{r}
#| code-fold: TRUE
#| output-fold: TRUE
arthritis_age_mon <- arthritis |>
  mutate(age_months = age * 12)

glimpse(arthritis_age_mon)
```

## Using `arrange` to sort numeric variable (age)

```{r}
#| code-fold: TRUE

# aesending order
arthritis |>
  arrange(age)
```

Quest: how about sorting age in a descending order?

```{r}
#| code-fold: TRUE
#| 
# descending order
arthritis |>
  arrange(desc(age))
```


## Creating a categorical variable: `cdai_level`

**Goal:** Create a new variable called `cdai_level` based on the CDAI data dictionary:

| Level | Condition |
|---|---|
| Remission | cdai ≤ 2.8 |
| Low Disease Activity | cdai > 2.8 and ≤ 10 |
| Moderate Disease Activity | cdai > 10 and ≤ 22 |
| High Disease Activity | cdai > 22 |

We use `case_when()` inside `mutate()` to assign categories:

```{r}
#| code-fold: TRUE
#| output-fold: TRUE
arthritis |>
  mutate(cdai_level = case_when(
    cdai <= 2.8             ~ "Remission",
    cdai > 2.8 & cdai <= 10 ~ "Low",
    cdai > 10  & cdai <= 22 ~ "Moderate",
    cdai > 22               ~ "High"))
```
 
To save the result back into `arthritis`:

```{r}
#| code-fold: TRUE
arthritis <- arthritis |>
  mutate(cdai_level = case_when(
    cdai <= 2.8             ~ "Remission",
    cdai > 2.8 & cdai <= 10 ~ "Low",
    cdai > 10  & cdai <= 22 ~ "Moderate",
    cdai > 22               ~ "High"))

glimpse(arthritis)
```

## Converting variable types

**Goal:** 

1. Make `cdai_level` a factor and `age` an integer.
2. Manually define the order of `cdai_level` to "High", "Moderate", "Low", "Remission".

```{r}
#| code-fold: TRUE
arthritis |>
  mutate(
    cdai_level = factor(cdai_level,levels = c("High", "Moderate", "Low", "Remission")),
    age = as.integer(age))
```

### Use `arrange()` sorts factors vs. characters:

```{r}
#| code-fold: TRUE
#| output-fold: TRUE
#| 
# Without levels: alphabetical (High, Low, Moderate, Remission)
arthritis |>
  arrange(cdai_level)

# With levels: manually defined order
arthritis |>
  mutate(cdai_level = factor(cdai_level,
    levels = c("Remission", "Low", "Moderate", "High"))) |>
  arrange(cdai_level)
```


Common type-conversion functions:

| Function | Result |
|---|---|
| `as.factor()` | factor |
| `as.numeric()` | numeric |
| `as.integer()` | integer |
| `as.double()` | double |
| `as.character()` | character |
 

## Your Turn

Using the `arthritis` dataset, complete the two tasks below.

### Task 1: Categorize years since diagnosis

Create a new variable called `yrs_category` that groups `yrs_from_dx` into three stages:

| Category | Condition |
|---|---|
| Early | yrs_from_dx ≤ 5 |
| Established | yrs_from_dx > 5 and ≤ 15 |
| Long-standing | yrs_from_dx > 15 |

Then make it a factor with levels ordered from earliest to longest: `"Early"`, `"Established"`, `"Long-standing"`. 

```{r}
#| code-fold: TRUE
arthritis |>
  mutate(
    yrs_category = case_when(
      yrs_from_dx <= 5                        ~ "Early",
      yrs_from_dx > 5  & yrs_from_dx <= 15   ~ "Established",
      yrs_from_dx > 15                        ~ "Long-standing"),
    yrs_category = factor(yrs_category, levels = c("Early", "Established", "Long-standing")))
```

### Task 2: Sort by disease stage and age

Extend your pipeline from Task 1: sort the result first by `yrs_category` (earliest to longest), then by `age` in descending order within each stage.

```{r}
#| code-fold: TRUE
arthritis |>
  mutate(
    yrs_category = case_when(
      yrs_from_dx <= 5                        ~ "Early",
      yrs_from_dx > 5  & yrs_from_dx <= 15    ~ "Established",
      yrs_from_dx > 15                        ~ "Long-standing"),
    yrs_category = factor(yrs_category, levels = c("Early", "Established", "Long-standing"))) |>
  arrange(yrs_category, desc(age)) 
```

