---
title: "Importing Data in R"
author: "Yang Meng"
output: 
  xaringan::moon_reader:
    css: ["slide-style.css", "https://cdnjs.cloudflare.com/ajax/libs/font-awesome/5.15.3/css/all.min.css"]
    lib_dir: libs
    seal: false
    nature:
      beforeInit: "cols_macro.js"
      ratio: 16:9
      highlightStyle: "pygments"
      highlightLines: true
      highlightLanguage: "r"

---

class: title-slide


```{r echo = FALSE, message = FALSE}
library(tidyverse)
options(scipen = 999)
```


<br>
<br>
.right-panel[ 

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$author`

Adapted from slides by Mine Dogucu
]


---

## Importing .csv Data 

`.csv` stands for comma separated values

```{r eval = FALSE}
readr::read_csv("dataset.csv")
```


---

## Importing Excel Data

```{r eval = FALSE}
readxl::read_excel("dataset.xlsx")
```



---
## Importing Excel Data

```{r eval = FALSE}
readxl::read_excel("dataset.xlsx", sheet = 2)
```

---

## Importing SAS, SPSS, Stata Data

```{r eval = FALSE}
library(haven)
# SAS
read_sas("dataset.sas7bdat")
# SPSS
read_sav("dataset.sav")
# Stata
read_dta("dataset.dta")
```

---

## Writing data files

```{r eval = FALSE}
write_csv(data_frame, "dataset.csv")
write_xls(data_frame, "dataset.xlsx")
```

Similar code pattern holds true for other data formats.

---

## Practice

- Open an `.Rmd` file.
- Download and import Alzheimer dataset. 
- Summarize information about the age and education of participants in the Alzheimer dataset.


---

## Practice 

1. Change the name of the variable female to gender.

2. Recode level 0 as `"male"` and level 1 as `"female"`

3. Count these levels. 

4. Write the summary table with counts into a `.csv` file.

---

## Practice

Download the HIV Antibody Test from the [CDC website](https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=Laboratory&CycleBeginYear=2017). Import it into R. 
Note that this file has an `.XPT` extension which we have not seen before. 
Let's see if we can open it using one of the packages we have already seen. 



