---
title: "Lab-01a-practice"
author: "Sara Tyo"
output: html_document
---

Follow Along with slides for Lab-01a-practice


See the the grey chunk below?

```{r}
# This is a comment
# comments are made using hashtags '#'
# comments don't do anything when you run the chunk
```

These  are called 'code chunks', that's where you can run R code by hitting the green right triangle at the top-right of a code chunk (you can also hit ctrl+enter/ctrl+shift+enter).

To make a new code chunk you can select the +C in the toolbar, to the left of the 'run' button, or hit ctrl+alt+i

## Panes

Look, There are 4 main panels. 

We primarily work here, in the upper left, where we write/type out code.

## Creating Objects: Assignment Operator

```{r}
birth_year <- 2000 # this is my birth year, maybe fill in your own!

birth_year
```

- Birth Year is the object being assigned to.

## Case Sensitive

```{r Wrong}
my_age <- 2024 - birth_year
My_age
```

Why is it throwing an error?

Solve the issue in this chunk and try to print the value of your age
```{r}
my_age
```

## Getting Help

In any code chunk the '?' to the left of an item/function can help to know what a function does.

```{r}
?mean
```

- Brief Description

- How function is used

- What arguments it takes

- Even provides examples

Look up what the function 'plot' does either using '?plot' or the help search bar. 

## R Variable Types

```{r}
class("charater")             # character
typeof("charater")            # character

class(3L)                     # integer
typeof(3L)                    # integer

class(3.0)                    # numeric
typeof(3.0)                   # double

class(as.factor(c("male","female"))) # factor
as.factor(c("male","female"))        # factor

class(TRUE)                   # logical
typeof(TRUE)                  # logical
```


## Some basic data structures

- variable: a single item 
- vector: multiple variables
- matrix: multiple variables with some structure/dimension

```{r}
temp_variable <- 1
temp_vector <- c(1, 2, 3)
temp_matrix <- matrix(c(1,2,3,4), nrow=2, ncol=2, 
                     byrow=TRUE)

temp_variable
temp_vector
temp_matrix
```

# Simple Math

```{r}
3+3 # add 3+3=6
4-1 # subtract 4-1=3

3/3 # divide 3/3=1
4*1 # multiply 4*1=4
```

Try and write your own to solve:

$$\frac{e^4*|-ln(3)|}{-19*45}$$

```{r}
# write code here
```


## For Loop

Remember this is how we generally write For Loops

if(condition){
   Do Something
} else {
   Do Something Else
}

```{r Example For Loop}
height <- 63
if(height == 63){
  print("Fantastic! Enjoy the rollercoaster")
} else {
  print("OOPS! Sorry, you'll have to wait until next year")
}

money <- 10
greed <- 1
if(money-1000 < 0 | greed > 50){
  print("Sorry I don't have enough money to loan you")
}else{
  print("Sure, here's 1000$.")
}
```

Try and play around with the amount assigned to 'money' and 'greed'.

Now, Try and make your own for loop!

```{r}
if(){
  print("")
}else{
  print("")
}
```

## Creating Functions

We can also make functions in R. 

Functions are useful when we find ourselves rewriting the same code over and over again. 

Functions are made using the following syntax, where it's important to prespecify the input arguments, and how you want them processed.

```{r}
my_function <- function(input){
  if(input == TRUE){
    return("Fantastic!")
  }
  else{
    return("OOPS!")
  }
}
```


- What does this function take in, what are the arguments?

- What does this function do with those arguments?

Try and get it to to return both results.

```{r}

```

