---
title: "Improving Data Visualization"
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    number-sections: true
    embed-resources: true
---

# Import data

```{r } 
#| message: false
#| code-fold: TRUE

library(openintro)
library(tidyverse)
library(ggplot2)
babies <- babies %>% 
  mutate(smoke = as.logical(smoke),
         parity = as.logical(parity))
 

```

Read documentation for this data set. Try `??babies`


# Scatter plot

Task:

- Using the `babies` data,   
- Map `gestation` to x-axis, `bwt` to y-axis, `smoke` to shape and color.
- Add a layer of points and set the size of the points to 4.


```{r } 
#| message: false
#| code-fold: TRUE
#| fig-width: 6
#| fig-height: 4
ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point(size = 4) 
```


# Add labels to axis using `labs()`

Task: 

- Using the `babies` data,   
- Map `gestation` to x-axis, `bwt` to y-axis, `smoke` to shape.
- Add a layer of points and set the size of the points to 4.
- [Add labels to x-axis (Gestation), y-axis (Birth Weight), and the title of the plot (Palmer babies).]{.highlight-text}

```{r } 
#| message: false
#| code-fold: TRUE
#| fig-width: 6
#| fig-height: 4

ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point(size = 4) +
  labs(x = "Gestation (days)", #<<
       y = "Birth Weight (ounces)", #<<
       title = "Baby Gestation and Birth Weight by Mother's Smoker Status") #<<
```

# Themes

## Theme options:

Let's customize the theme. Try add a layer of `theme_bw()`.

Other theme options: 

- `theme_gray()`
- `theme_bw()`
- `theme_classic()`
- `theme_dark()`
- `theme_minimal()`
- `theme_void()`


```{r warning=FALSE } 
#| message: false
#| code-fold: TRUE
#| fig-width: 6
#| fig-height: 4

ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point() +
labs(x = "Gestation (days)",
       y = "Birth Weight (ounces)", 
       title = "Gestation and Birth Weight by Mother's Smoker Status") +
  theme_bw() #<<
```



```{r warning=FALSE } 
#| message: false
#| code-fold: TRUE
#| output-fold: TRUE
#| fig-width: 4
#| fig-height: 3
ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point() +
  labs(x = "Gestation (days)",
       y = "Birth Weight (ounces)", 
       title = "Baby Gestation and Birth Weight by Mother's Smoker Status") +
  theme_gray()
```


```{r warning=FALSE } 
#| message: false
#| code-fold: TRUE
#| output-fold: TRUE
#| fig-width: 4
#| fig-height: 3
ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point() +
  labs(x = "Gestation (days)",
       y = "Birth Weight (ounces)", 
       title = "Baby Gestation and Birth Weight by Mother's Smoker Status") +
  theme_classic()
```



```{r warning=FALSE } 
#| message: false
#| code-fold: TRUE
#| output-fold: TRUE
#| fig-width: 4
#| fig-height: 3
ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point() +
  labs(x = "Gestation (days)",
       y = "Birth Weight (ounces)", 
       title = "Baby Gestation and Birth Weight by Mother's Smoker Status") +
  theme_dark()
```


```{r  warning=FALSE} 
#| message: false
#| code-fold: TRUE
#| output-fold: TRUE
#| fig-width: 4
#| fig-height: 3
ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point() +
  labs(x = "Gestation (days)",
       y = "Birth Weight (ounces)", 
       title = "Baby Gestation and Birth Weight by Mother's Smoker Status") +
  theme_minimal()
```


```{r warning=FALSE} 
#| message: false
#| code-fold: TRUE
#| output-fold: TRUE
#| fig-width: 4
#| fig-height: 3
ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point() +
  labs(x = "Gestation (days)",
       y = "Birth Weight (ounces)", 
       title = "Baby Gestation and Birth Weight by Mother's Smoker Status") +
  theme_void()
```
 

## Explore `theme()`


```{r}
?theme
```

There are more aspects of a theme that we can control. 


```{r  warning=FALSE} 
#| message: false
#| code-fold: TRUE
#| output-fold: TRUE
#| out-width: "80%"
ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point() +
  labs(x = "Gestation (days)",
       y = "Birth Weight (ounces)", 
       title = "Baby Gestation and Birth Weight by Mother's Smoker Status") +
  theme_bw() +
  theme(title = 
          element_text(size = 12),
        axis.title = 
          element_text(size = 10,
                       face="bold"))
  
```
 
 
# Layout panels in a grid 

`facet_grid()` splits your plot into a grid of panels — one panel per group — so you can compare patterns across categories side by side. The formula `rows ~ cols` controls the direction: `.~var` places panels side by side in columns, while `var~.` stacks them in rows.
 
 
```{r } 
#| message: false
#| code-fold: TRUE
#| output-fold: TRUE
#| out-width: "80%"
ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point(size = 4) +
  facet_grid(.~parity) #<<
```

 
```{r } 
#| message: false
#| code-fold: TRUE
#| output-fold: TRUE
#| out-width: "80%"
ggplot(babies,
       aes(x = gestation,
           y = bwt,
           shape = smoke,
           color = smoke)) +
  geom_point(size = 4) +
  facet_grid(parity~.) #<<
```
 
# Summary

You can do much more in ggplot. 

- Use images 
- Make maps
- Pick colors that you want, 

Resources:

- [the ggplot flipbook](https://evamaerey.github.io/ggplot_flipbook/ggplot_flipbook_xaringan.html#1) 
- [coloring for colorblindness](https://davidmathlogic.com/colorblind/#%23D81B60-%231E88E5-%23FFC107-%23004D40)
- [Gallery](https://r-graph-gallery.com/ggplot2-package.html)
- [BBC](https://bbc.github.io/rcookbook/)   
- [fivethirtyeight](https://blog.revolutionanalytics.com/2016/07/data-journalism-with-r-at-538.html?utm_content=bufferc03e0&utm_medium=social&utm_source=twitter.com&utm_campaign=buffer)  
- [NY Times](https://livefreeordichotomize.com/2021/04/07/nytimes-map-how-to/)


# Practice

Can you make this plot?

## Relationship between smoke mother v.s. birth weight

**Hint**: use `filter()` and `is.na()` to remove the NA values in the smoke column


```{r warning=FALSE}
#| message: false
#| code-fold: TRUE
#| output-fold: TRUE
#| out-width: "80%"

babies %>% 
  filter(!is.na(smoke)) %>% 
  ggplot(
       aes(x = smoke,
           y = bwt)) +
  geom_violin() +
  labs(x = 'Mother\'s smoke history', y = 'Birth weight')+
  theme_classic()
```


## Relationship between mother's age group v.s. birth weight
**Hint**: use `case_when()` to create a new variable `age_group` and use `geom_jitter()` and `geom_boxplot()` to make this figure!

```{r warning=FALSE}
#| message: false
#| code-fold: TRUE
#| output-fold: TRUE
#| out-width: "80%"
babies %>% 
  mutate(
    age_grp = case_when(
      age >= 30 ~ '30+',
      20 <= age & age < 30 ~ '20-30',
      age < 20 ~ 'under 20'
    ),
    age_grp = factor(age_grp, levels = c('under 20','20-30','30+'))
  ) %>% 
  filter(!is.na(age_grp)) %>% 
  ggplot(aes(
    x = age_grp,
           y = bwt,
           shape = age_grp,
           color = age_grp)) +
  geom_jitter(width = 0.2,alpha = 0.3) +
  geom_boxplot(alpha = 0.5, size = 1) +
  labs(x = 'Mother\'s age group', y = 'Birth weight')+
  theme_classic()
```

