---
title: "Introduction to tidyverse"
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    number-sections: true
    embed-resources: true
---

```{r}
#| echo: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## The tidyverse Style Guide

> canyoureadthissentence?

::: {callout-important}
Good coding style is like correct punctuation: you can manage without it, but it makes things easier to read.
:::

The most important thing about the tidyverse style guide is that it provides **consistency**, making code easier to write because you need to make fewer decisions.

::: {.callout-tip collapse="false"}
See: [The tidyverse style guide](https://style.tidyverse.org/)
:::

::: {.callout-note collapse="false" icon="false"}
### File Names

Avoid special characters in file names — stick with numbers, letters, -, and \_.

```{r}
#| eval: false

# Good
fit_models.R
utility_functions.R

# Bad
fit models.R
foo.r
stuff.r
lab1.r
```
:::

::: {.callout-note collapse="true" icon="false"}
### Syntax

- Leave spaces before and after operators (e.g. \<-, =).
- Put a space after a comma, not before.
- Object names should be all lowercase, with words separated by an underscore.

```{r}
#| eval: false

#---------Spacing ------------
# Good
age <- c(6, 9, 15)
# Bad
age <- c(6,9,15)

# Good
data.frame(age_kid = age)
# Bad
data.frame(Age_kid=age)

#---------Object name ------------
# Good
day_one <- 1
day_1 <- 1
# Bad
DayOne <- 1
first_day_of_the_month <- 1

#---------Parentheses ------------
# Good
mean(x, na.rm = TRUE)
# Bad
mean (x, na.rm = TRUE)
mean( x, na.rm = TRUE )


#---------Function------------
# Good
function(x) {}

# Bad
function (x) {}
function(x){}
```
:::



## The Pipe Operator

**What is the average of 4, 8, 16 — approximately?**

Breaking the problem down:

1.  What is the average of **4, 8, 16** approximately?
2.  What is the **average** of 4, 8, 16 approximately?
3.  What is the average of 4, 8, 16 **approximately**?

::: {.callout-note collapse="false"}
### Solution 1: Functions within Functions

```{r}
#| code-fold: TRUE
#| output-fold: TRUE
c(4, 8, 16)
```

```{r}
#| code-fold: TRUE
#| output-fold: TRUE
mean(c(4, 8, 16))
```

```{r}
#| code-fold: TRUE
#| output-fold: TRUE
round(mean(c(4, 8, 16)))
```

**Problem:** Things get messy and harder to read as operations become more complex.
:::

::: {.callout-note collapse="true"}
### Solution 2: Creating Objects

```{r}
#| code-fold: TRUE
#| output-fold: TRUE
numbers <- c(4, 8, 16)
numbers
```

```{r}
#| code-fold: TRUE
#| output-fold: TRUE
avg_number <- mean(numbers)
avg_number
```

```{r}
#| code-fold: TRUE
#| output-fold: TRUE
round(avg_number)
```

**Problem:** We end up with too many objects cluttering the `Environment`.
:::

:::::: {.callout-note collapse="true"}
### Solution 3: The Pipe Operator `|>`

First, load the tidyverse:

```{r}
#| message: false
#| code-fold: TRUE
#| output-fold: TRUE

# install.packages("tidyverse")
library(tidyverse)
```

The pipe `|>` passes the output of one function as the first argument of the next.

**Keyboard shortcut:** `Ctrl/Cmd + Shift + M`

```{r}
#| code-fold: TRUE
#| output-fold: TRUE
c(4, 8, 16) |>
  mean() |>
  round()
```

Read this as: Combine 4, 8, and 16, **then** take the mean, **then** round the result.

This is equivalent to the composite function $$f \circ g \circ h(x)$$, or `round(mean(c(4, 8, 16)))` — but much easier to read.

::::: columns
::: {.column width="50%"}
**Generalized form**

```{r}
#| eval: false
h(x) |>
  g() |>
  f()
```
:::

::: {.column width="50%"}
**Our example**

```{r}
#| eval: false
#| code-fold: TRUE
#| output-fold: TRUE
c(4, 8, 16) |>
  mean() |>
  round()
```
:::
:::::
::::::



## Practice: The Pipe Operator

### Exercise 1: Top 3 Scores

A class got these quiz scores: `72, 88, 95, 61, 79, 100, 83`. Use the pipe to find the **top 3 scores**, from highest to lowest.

**Hint:** try `sort()`, `rev()`, and `head()`.

```{r}
#| code-fold: TRUE
c(72, 88, 95, 61, 79, 100, 83) |>
  sort() |>
  rev() |>
  head(3)
```

### Exercise 2: How Many Words?

Count the number of words in the sentence `"To be or not to be that is the question"`.

**Hint:** try `strsplit()`, `unlist()`, and `length()`.

```{r}
#| code-fold: TRUE
"To be or not to be that is the question" |>
  strsplit(split = " ") |>
  unlist() |>
  length()
```
