---
title: "Changing Variables"
author: "Dr. Mine Dogucu"
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    number-sections: true
    embed-resources: true
---

## Loading Data
 
```{r}
#| message: false
#| code-fold: TRUE

library(tidyverse)
options(scipen = 999)

arthritis <- read_csv("https://raw.githubusercontent.com/cosmos-uci-dshs/data/main/RheumArth_Tx_AgeComparisons.csv") |>
  janitor::clean_names() |>
  mutate(sex = case_when(sex == 0 ~ "female",
                         sex == 1 ~ "male"),
         sex = as.factor(sex),
         age_gp = case_when(age_gp == 1 ~ "control",
                            age_gp == 2 ~ "elderly"),
         age_gp = as.factor(age_gp),
         cdai_yn = case_when(cdai_yn == 1 ~ "no",
                             cdai_yn == 2 ~ "yes"),
         cdai_yn = as.factor(cdai_yn)) |>
  select(age, age_gp, sex, yrs_from_dx, cdai)
```
 

## Data Overview

We are working with the `arthritis` dataset. Here is a quick look at its structure:

```{r}
glimpse(arthritis)
```

---

## Mutating Variables

We use `mutate()` to create or modify variables in a data frame.


## Creating a numeric variable: `age_months`

**Goal:** Create a new variable called `age_months` that represents `age` in months.

```{r}
#| code-fold: TRUE
arthritis |>
  mutate(age_months = age * 12)
```

Note that this does not change the original `arthritis` object. To save the result, assign it:

```{r}
#| code-fold: TRUE
#| output-fold: TRUE
arthritis_age_mon <- arthritis |>
  mutate(age_months = age * 12)

glimpse(arthritis_age_mon)
```

## Creating a categorical variable: `cdai_level`

**Goal:** Create a new variable called `cdai_level` based on the CDAI data dictionary:

| Level | Condition |
|---|---|
| Remission | cdai ≤ 2.8 |
| Low Disease Activity | cdai > 2.8 and ≤ 10 |
| Moderate Disease Activity | cdai > 10 and ≤ 22 |
| High Disease Activity | cdai > 22 |

We use `case_when()` inside `mutate()` to assign categories:

```{r}
#| code-fold: TRUE
#| output-fold: TRUE
arthritis |>
  mutate(cdai_level = case_when(
    cdai <= 2.8             ~ "Remission",
    cdai > 2.8 & cdai <= 10 ~ "Low",
    cdai > 10  & cdai <= 22 ~ "Moderate",
    cdai > 22               ~ "High"))
```

We can pipe directly into `ggplot()` to visualize the new variable:

```{r}
#| code-fold: TRUE 
arthritis |>
  mutate(cdai_level = case_when(
    cdai <= 2.8             ~ "Remission",
    cdai > 2.8 & cdai <= 10 ~ "Low",
    cdai > 10  & cdai <= 22 ~ "Moderate",
    cdai > 22               ~ "High")) |>
  ggplot(aes(x = cdai_level)) +
  geom_bar()
```

To save the result back into `arthritis`:

```{r}
#| code-fold: TRUE
arthritis <- arthritis |>
  mutate(cdai_level = case_when(
    cdai <= 2.8             ~ "Remission",
    cdai > 2.8 & cdai <= 10 ~ "Low",
    cdai > 10  & cdai <= 22 ~ "Moderate",
    cdai > 22               ~ "High"))

glimpse(arthritis)
```

## Converting variable types

**Goal:** Make `cdai_level` a factor and `age` an integer.

```{r}
#| code-fold: TRUE
arthritis |>
  mutate(cdai_level = as.factor(cdai_level),
         age = as.integer(age))
```

Common type-conversion functions:

| Function | Result |
|---|---|
| `as.factor()` | factor |
| `as.numeric()` | numeric |
| `as.integer()` | integer |
| `as.double()` | double |
| `as.character()` | character |
 
