---
title: "Lab-01b-practice"
author: "Sara Tyo"
output: html_document
---

Follow Along with slides for Lab-01b-practice

## R packages

Try running the following code:

```{r}
beep()
```

Why are we seeing this error?

Answer: We don't have this package downloaded.

Think of packages as items, when we have an error it means we have yet to get/purchase an item.

## Installing 'Getting' a Package

You can copy or type install.packages("beepr") in the console, or in the chunk below.

NOTE: we only ever need to install a package once. Once downloaded, we can 'call' it and use it whenever we want.

```{r}
install.packages("beepr")
```

We can also use the 'packages' search in the bottom-right pane

## Using Packages and Functions from Packages 

- We want to use the beep() function from the beepr package. So we first need to call/open the package.
- This is like getting the object we purchased in order to use it.

How do we 'get it'?

We use the function library() to recall and load in the package.

```{r}
library("beepr")
```

Now try and run beep()

```{r}
beep()
```

YAY!

What if we want a different sound? Where do we go again if we don't know what to do, or maybe want to learn more about a particular function?

Use help search bar, or '?'

```{r Help using ?}
?beep
```

- What kinds of sounds are there from the description?

- 1. ping
- 2...
- 3...

- Try out different sound types

```{r}
beep(sound="ping")
```

## More Fun

Now, use the say() function from the library cowsay and see what it doesomk0    

```{r}
install.packages("cowsay")
library(cowsay)
```

Need Help?
```{r}
?say()
```

- It will Print out default "Hello world", and generate an image of a cat
- we can change what it says by adjusting the 'what' argument
- we can change the animal by adjusting the 'by' argument

What are some of the animals it can generate?
-

Are there other things you can change?

Play around below!

```{r Have Fun}
say(what="Have Fun!", by="cat", what_color="green")
```

