---
title: "Controlling Model Complexity"
subtitle: "Dimensionality Reduction and Regularization"
author: "Babak Shahbaba"
execute:
  echo: true
  warning: false
  message: false
format:
  revealjs:
    title-slide: true
    theme: default
    chalkboard: true
    scrollable: true
    self-contained: false
    highlight-style: pygments
    code-line-numbers: true
    code-overflow: wrap
    slide-number: true
    transition: fade
    progress: true
    controls: false
    ratio: 3:2
    code-fold: true
    footer: "Controlling Complexity • COSMOS"
---

```{r}
#| echo: false
#| error: false

library(tidyverse)
library(ggrepel)
library(glmnet)
library(pROC)

theme_set(theme_minimal(base_size = 18))

set.seed(2026)
```

## Learning objectives

This lecture studies how model complexity affects prediction and interpretation.

By the end, you should be able to:

- explain why a model can fit training data well but predict poorly;
- describe principal component analysis as a change of coordinates;
- explain how ridge and lasso control coefficient size;
- use cross-validation to select the strength of regularization; and
- compare binary classifiers using ROC curves and AUC.

---

## The central statistical problem

Suppose a model has many predictors or highly flexible functions.

- It can represent complicated relationships.
- It can also fit random noise as though it were signal.

The goal is not to minimize training error at any cost. The goal is to minimize error on **new observations**.

::: {.callout-important}
Complexity control deliberately restricts a model so that it generalizes more reliably.
:::

---

## Training error and test error

```{r}
#| echo: false
#| fig-height: 5.4

complexity_curve <- tibble(
  complexity = seq(0, 10, length.out = 300),
  training = 1.8*exp(-0.28*complexity) + 0.15,
  test = 0.55 + 0.055*(complexity - 5.2)^2
) %>%
  pivot_longer(-complexity, names_to = "error_type", values_to = "error")

ggplot(complexity_curve, aes(complexity, error, color = error_type)) +
  geom_line(linewidth = 1.4) +
  geom_vline(xintercept = 5.2, linetype = 2) +
  scale_color_manual(values = c(training = "#2C7FB8", test = "#D95F0E")) +
  labs(title = "Model complexity and predictive error",
       x = "Model complexity", y = "Error", color = NULL)
```

Training error typically falls with complexity. Test error often follows a U-shaped curve.

---

## Underfitting and overfitting

::: {.columns}
::: {.column width="48%"}
### Underfitting

- Model is too restricted.
- Important structure is missed.
- Training and test errors are both high.
- Associated with high bias.
:::

::: {.column width="48%"}
### Overfitting

- Model is excessively adaptive.
- Noise is treated as reproducible structure.
- Training error is low; test error is high.
- Associated with high variance.
:::
:::

The best predictive model usually lies between these extremes.

---

## Two strategies for complexity control

```{mermaid}
flowchart LR
  A["Many correlated predictors"] --> B["Principal components"]
  A --> C["Penalized regression"]
  B --> D["Use a small number of components"]
  C --> E["Shrink coefficients toward zero"]
```

---

## Principal component analysis

PCA creates new variables called **principal components**. Each component is a weighted combination of the original variables.

- PC1 follows the direction with the greatest variation.
- PC2 captures the greatest remaining variation while being perpendicular to PC1.
- Later components continue this process.

- **Loadings** describe how the original variables contribute to a component.
- **Scores** describe where each observation lies along that component.

---

## PCA as a rotation of coordinates

```{r}
#| echo: false
#| fig-height: 5.6

set.seed(18)
u <- rnorm(180, sd = 2.3)
v <- rnorm(180, sd = 0.45)
theta <- pi/5
rotated <- tibble(
  x1 = u*cos(theta) - v*sin(theta),
  x2 = u*sin(theta) + v*cos(theta)
)
pca_demo <- prcomp(rotated, center = TRUE, scale. = FALSE)
center <- colMeans(rotated)
dirs <- as.data.frame(pca_demo$rotation)

ggplot(rotated, aes(x1, x2)) +
  geom_point(alpha = 0.55) + coord_equal() +
  geom_segment(aes(x = center[1], y = center[2],
                   xend = center[1] + 3*dirs[1,1],
                   yend = center[2] + 3*dirs[2,1]),
               arrow = arrow(length = unit(0.15, "in")),
               linewidth = 1.3, color = "#2C7FB8") +
  geom_segment(aes(x = center[1], y = center[2],
                   xend = center[1] + 2*dirs[1,2],
                   yend = center[2] + 2*dirs[2,2]),
               arrow = arrow(length = unit(0.15, "in")),
               linewidth = 1.3, color = "#D95F0E") +
  labs(title = "Principal components define orthogonal directions",
       subtitle = "PC1 follows greatest variation; PC2 follows the remaining variation")
```

---

## Why standardization matters in PCA

Without standardization, PCA favors variables with large numerical variance.

- Temperature measured in degrees and population measured in millions have incomparable scales.
- Changing a variable from meters to millimeters can change unstandardized PCA.

Standardized PCA uses

$$
z_{ij}=\frac{x_{ij}-\bar x_j}{s_j},
$$


---

## Example: socioeconomic characteristics of states

R's `state.x77` dataset contains eight measurements for the 50 U.S. states.

```{r}
states <- as_tibble(state.x77, rownames = "state")
glimpse(states)
```

```{r}
state_x <- states %>% select(-state)
state_pca <- prcomp(state_x, center = TRUE, scale. = TRUE)

summary(state_pca)$importance[, 1:5]
```

The component signs are arbitrary: reversing every sign in one loading vector gives the same axis.

---

## Scree plot

```{r}
#| echo: false
#| fig-height: 5.4

state_var <- state_pca$sdev^2
scree <- tibble(
  component = seq_along(state_var),
  proportion = state_var/sum(state_var),
  cumulative = cumsum(proportion)
)

ggplot(scree, aes(component, proportion)) +
  geom_col(fill = "#2C7FB8") +
  geom_line(aes(y = cumulative), color = "#D95F0E", linewidth = 1.2) +
  geom_point(aes(y = cumulative), color = "#D95F0E", size = 2.5) +
  scale_x_continuous(breaks = 1:nrow(scree)) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Variance explained by the principal components",
       x = "Principal component", y = "Proportion / cumulative proportion")
```

A scree plot supports judgment; it does not supply a universally correct cutoff.


---

## PCA biplot

```{r}
#| echo: false
#| fig-height: 6

scores <- as_tibble(state_pca$x[, 1:2]) %>%
  mutate(state = states$state)

loading_scale <- 3.2
loading_plot <- as_tibble(state_pca$rotation[, 1:2], rownames = "variable") %>%
  mutate(across(c(PC1, PC2), ~ .x*loading_scale))

ggplot(scores, aes(PC1, PC2)) +
  geom_hline(yintercept = 0, color = "grey80") +
  geom_vline(xintercept = 0, color = "grey80") +
  geom_point(color = "#2C7FB8") +
  geom_text_repel(aes(label = state), size = 3, max.overlaps = 25) +
  geom_segment(data = loading_plot,
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               inherit.aes = FALSE, color = "#D95F0E",
               arrow = arrow(length = unit(0.12, "in"))) +
  geom_text_repel(data = loading_plot,
                  aes(PC1, PC2, label = variable),
                  inherit.aes = FALSE, color = "#D95F0E", size = 3.5) +
  labs(title = "PCA biplot for state characteristics")
```

---

## Principal component regression

Principal component regression (PCR) uses a selected number of component scores as predictors instead of using all the original variables directly.

The number $q$ controls complexity:

- small $q$: stronger dimension reduction;
- using all components: no dimensionality reduction.

Choose $q$ by cross-validation when prediction is the goal.

::: {.callout-note}
PCA preserves variation in the predictors, not necessarily the variation most useful for predicting the outcome. Cross-validation is therefore preferable to choosing $q$ from the scree plot alone.
:::

---

## Regularization retains the original predictors

Ordinary least squares chooses coefficients that fit the training outcomes as closely as possible. When predictors are numerous or highly correlated, those coefficients can become large and unstable.

Regularization still rewards good fit, but it also discourages large coefficients. This restriction can improve prediction on new data.

The tuning parameter $\lambda$ controls the strength of regularization:

- $\lambda=0$: ordinary least squares;
- larger $\lambda$: stronger shrinkage and a simpler model.

We choose $\lambda$ using cross-validation rather than training error.

---

## Ridge regression

Ridge regression minimizes

$$
\operatorname{RSS}(\beta)+\lambda\sum_{j=1}^{p}\beta_j^2.
$$

The squared-coefficient penalty shrinks every coefficient smoothly toward zero. 

Predictors are usually standardized first so the penalty treats variables measured in different units fairly.

---

## Ridge coefficient paths

```{r}
#| echo: false
#| fig-height: 5.8

set.seed(12)
n <- 100
p <- 7
latent <- rnorm(n)
X_sim <- sapply(1:p, function(j) 0.75*latent + rnorm(n, sd = 0.7))
colnames(X_sim) <- paste0("x", 1:p)
beta_true <- c(3, 2, 0, 0, -1.5, 0, 0.7)
y_sim <- as.vector(X_sim %*% beta_true + rnorm(n, sd = 2))

Xz <- scale(X_sim)
yz <- as.vector(scale(y_sim, center = TRUE, scale = FALSE))
lambdas <- 10^seq(-3, 3, length.out = 120)

ridge_coef <- map_dfr(lambdas, function(lambda) {
  b <- solve(crossprod(Xz) + lambda*diag(p), crossprod(Xz, yz))
  tibble(lambda = lambda, variable = colnames(Xz), coefficient = as.vector(b))
})

ggplot(ridge_coef, aes(lambda, coefficient, color = variable)) +
  geom_line(linewidth = 1) +
  scale_x_log10() +
  geom_hline(yintercept = 0, color = "grey65") +
  labs(title = "Ridge coefficient paths",
       x = expression(lambda), y = "Standardized coefficient")
```

Coefficients shrink smoothly toward zero but are not exactly zero.

---

## Lasso regression

The lasso minimizes

$$
\operatorname{RSS}(\beta)+\lambda\sum_{j=1}^{p}|\beta_j|.
$$

Unlike ridge, lasso can set some coefficient estimates exactly to zero.

This combines:

- **shrinkage:** reducing coefficient magnitudes;
- **selection:** excluding predictors from the fitted equation.

As $\lambda$ increases, smaller coefficients may reach zero and leave the model. This can create a simpler equation.

---

## Lasso coefficient paths

```{r}
#| echo: false
#| fig-height: 5.8

# Use the same simulated data as in the ridge path
lasso_path <- glmnet(
  Xz, yz, alpha = 1,
  standardize = FALSE, intercept = FALSE
)

lasso_coef <- as.data.frame(t(as.matrix(lasso_path$beta))) %>%
  mutate(lambda = lasso_path$lambda) %>%
  pivot_longer(-lambda, names_to = "variable",
               values_to = "coefficient")

ggplot(lasso_coef, aes(lambda, coefficient, color = variable)) +
  geom_line(linewidth = 1) +
  scale_x_log10() +
  geom_hline(yintercept = 0, color = "grey65") +
  labs(title = "Lasso coefficient paths",
       x = expression(lambda), y = "Standardized coefficient")
```

As $\lambda$ increases, lasso coefficients reach exactly zero at different points. This produces a sequence of increasingly sparse models.


---

## Selecting the regularization parameter

The value of $\lambda$ is a tuning parameter, not something estimated by minimizing training RSS.

- $\lambda=0$: least squares, least shrinkage.
- moderate $\lambda$: some bias, reduced variance.
- very large $\lambda$: coefficients approach zero, severe underfitting.

Use $K$-fold cross-validation:

1. divide training data into $K$ folds;
2. fit on $K-1$ folds across a grid of $\lambda$ values;
3. predict the held-out fold;
4. average validation error across folds;
5. select $\lambda$, then refit using all training data.

---

## Cross-validation is part of model fitting

```{mermaid}
flowchart LR
  A["Training data"] --> B["Cross-validation folds"]
  B --> C["Select λ or q"]
  C --> D["Refit on all training data"]
  D --> E["Evaluate once on test data"]
```

::: {.callout-warning}
If the test set is repeatedly used to choose $q$, $\lambda$, transformations, or predictors, it becomes part of the training process and no longer provides an honest final evaluation.
:::

---

## Binary classification example

We simulate a binary outcome with a deliberately challenging predictor set:

- 400 observations;
- 20 useful but highly correlated measurements;
- 40 unrelated noise variables;
- overlapping classes, so perfect prediction is impossible.

```{r}
set.seed(2)
n <- 400
z1 <- rnorm(n)
z2 <- rnorm(n)

# Ten noisy measurements of each underlying signal
signal_1 <- sapply(1:10, function(j) z1 + rnorm(n, sd = 0.7))
signal_2 <- sapply(1:10, function(j) z2 + rnorm(n, sd = 0.7))
noise <- matrix(rnorm(n * 40), nrow = n)

x <- cbind(signal_1, signal_2, noise)
colnames(x) <- paste0("x", 1:60)

probability <- plogis(0.7*z1 - 0.7*z2)
y <- factor(rbinom(n, 1, probability), levels = c(0, 1))

train_id <- sample(seq_len(n), 200)
x_train_raw <- x[train_id, ]
x_test_raw <- x[-train_id, ]
y_train <- y[train_id]
y_test <- y[-train_id]
```

The same 200 training and 200 test observations are used for all three models.

---

## Three logistic classification models

We compare:

1. **Logistic regression:** estimates 60 predictor coefficients from only 200 training observations.
2. **PCA–logistic regression:** summarizes the correlated predictors using two principal components, then fits logistic regression.
3. **Lasso logistic regression:** considers all 60 predictors but shrinks less useful coefficients toward zero.

---

## Prepare predictors and principal components

Standardization and PCA are learned from the **training set only** and then applied to the test set.

```{r}
train_mean <- colMeans(x_train_raw)
train_sd   <- apply(x_train_raw, 2, sd)
x_train <- scale(x_train_raw, center = train_mean, scale = train_sd)
x_test  <- scale(x_test_raw,  center = train_mean, scale = train_sd)

pca_fit <- prcomp(x_train, center = FALSE, scale. = FALSE)
variance_explained <- pca_fit$sdev^2 / sum(pca_fit$sdev^2)
cumsum(variance_explained)[1:5]
```


---

## Fit the three models in R

```{r}
# Ordinary logistic regression
logit_data <- data.frame(y = y_train, x_train)
logit_fit <- glm(y ~ ., data = logit_data, family = binomial)

# Logistic regression using the first two principal components
pc_train <- as.data.frame(pca_fit$x[, 1:2, drop = FALSE])
pc_test <- as.data.frame(
  predict(pca_fit, newdata = x_test)[, 1:2, drop = FALSE]
)
pc_train$y <- y_train
pca_logit_fit <- glm(y ~ ., data = pc_train, family = binomial)

# Lasso logistic regression; lambda is chosen by cross-validation
set.seed(2)
lasso_fit <- cv.glmnet(
  x_train, y_train,
  family = "binomial", alpha = 1,
  type.measure = "auc", nfolds = 10
)
```

---

## Test-set predicted probabilities

```{r}
prob_logit <- predict(
  logit_fit,
  newdata = data.frame(x_test), type = "response"
)

prob_pca <- predict(
  pca_logit_fit,
  newdata = pc_test, type = "response"
)

prob_lasso <- as.numeric(
  predict(lasso_fit, newx = x_test,
          s = "lambda.min", type = "response")
)

selected_lasso <- coef(lasso_fit, s = "lambda.min")

```


---

## Confusion matrix

After choosing a probability threshold, each prediction is classified as positive or negative. Comparing predictions with the true outcomes gives a **confusion matrix**.

| | **Actually positive** | **Actually negative** |
|---|---:|---:|
| **Predicted positive** | True positive (TP) | False positive (FP) |
| **Predicted negative** | False negative (FN) | True negative (TN) |

- **Sensitivity** asks: among the actual positive cases, what proportion did we detect?

$$
\text{Sensitivity}=\frac{TP}{TP+FN}
$$

- **Specificity** asks: among the actual negative cases, what proportion did we correctly reject?

$$
\text{Specificity}=\frac{TN}{TN+FP}
$$

Changing the probability threshold changes the entries of the confusion matrix and therefore changes sensitivity and specificity.

---

## ROC curves compare discrimination

A classifier first produces a score or probability. We then choose a threshold above which an observation is classified as positive.

Changing the threshold changes two quantities:

- **Sensitivity (true-positive rate):** the proportion of positive cases correctly identified.
- **False-positive rate:** the proportion of negative cases incorrectly classified as positive; this equals $1-\text{specificity}$.

An **ROC curve** plots sensitivity against the false-positive rate across all possible thresholds.

- Curves closer to the upper-left corner indicate better discrimination.
- The diagonal line represents performance similar to random guessing.
- **AUC** summarizes the curve: $0.5$ indicates little discrimination and $1$ indicates perfect discrimination.

---

## ROC comparison of the three models

```{r}
#| echo: false
#| fig-height: 5.5

roc_models <- list(
  "Logistic regression" = roc(y_test, prob_logit, quiet = TRUE),
  "PCA + logistic regression" = roc(y_test, prob_pca, quiet = TRUE),
  "Lasso logistic regression" = roc(y_test, prob_lasso, quiet = TRUE)
)

auc_values <- map_dbl(roc_models, ~ as.numeric(auc(.x)))
names(roc_models) <- paste0(
  names(roc_models), " (AUC = ", round(auc_values, 3), ")"
)

ggroc(roc_models, legacy.axes = TRUE, linewidth = 1.2) +
  geom_abline(intercept = 0, slope = 1, linetype = 2,
              color = "grey55") +
  coord_equal() +
  labs(
    title = "Test-set ROC curves",
    x = "False-positive rate (1 − specificity)",
    y = "True-positive rate (sensitivity)",
    color = NULL
  )
```

---

## Interpreting ROC curves and AUC

- Curves closer to the upper-left corner show better discrimination.
- The diagonal line represents performance similar to random guessing.
- **AUC** summarizes the curve: $0.5$ indicates little discrimination and $1$ indicates perfect discrimination.
- ROC curves compare models across thresholds; they do not identify the best clinical threshold automatically.


---

## Interpreting ROC curves and AUC

```{r}
tibble(
  model = c("Logistic regression",
            "PCA + logistic regression",
            "Lasso logistic regression"),
  test_auc = round(auc_values, 3)
) %>%
  arrange(desc(test_auc))
```

Small AUC differences on one test split should not be overinterpreted. Repeated resampling provides a more stable comparison.

