---
title: "Parameter Estimation Practice in R"
author: "Brian Schetzsle"
output: 
  xaringan::moon_reader:
    css: ["slide-style.css"]
    lib_dir: libs
    seal: false
    nature:
      ratio: 16:9
      highlightStyle: "pygments"
      highlightLines: true
      highlightLanguage: "r"

---

class: title-slide

```{r echo = FALSE}
library(fabricerin)
```

<br>
<br>
.right-panel[ 
<br>

# `r rmarkdown::metadata$title`

### `r rmarkdown::metadata$author`

]

---

class: middle

## Reminder

- Close all apps on your computer other than zoom.  
- Open slides for this session from the cluster website (https://uci-dshs.netlify.app/). 
- We will be working with the capstone project [data](https://github.com/COSMOS-DataScience/slides/blob/main/data/alzheimer_data.csv).

---

## Conditional Statements in R

```{r eval = FALSE}
if(condition){
  print("Fantastic!")
} else {
  print("OOPS!")
}
```

```{r eval = FALSE}
ifelse(condition, yes = "Fantastic!", no = "OOPS!" )
```

```{r eval = FALSE}
case_when(
  condition ~ "Fantastic!",
  TRUE ~ "OOPS!"
)
```

---

## Loops

```{r eval = FALSE}
for(i in 1:10){
  print(i)
}
```

```{r eval = FALSE}
i <- 1
while(i <= 10){
  print(i)
  i <- i + 1
}
```

```{r eval = FALSE}
apply(data, MARGIN = 2, FUN = mean)
```

---

## Creating Functions

```{r eval = FALSE}
my_function <- function(input){
  if(input == TRUE){
    return("Fantastic!")
  }
  else{
    return("OOPS!")
  }
}
```

---

## Loading the Data

-- Now let's get into statistical inference and hypothesis testing

```{r warning=FALSE, message=FALSE}
library(tidyverse)

data <- read_csv("https://raw.githubusercontent.com/COSMOS-DataScience/slides/main/data/alzheimer_data.csv")

attach(data)

n = nrow(data)
```

---

## Loading the Data

```{r}
glimpse(data)
```

---

## Estimating Popluation Mean of a Continuous Variable

- Estimate the height of older American adults using the Alzheimer's data.

```{r}
mean(height)
```
- The math behind this code:

$$
\overline{\text{height}} = \frac{1}{n} \sum_{i=1}^n \text{height}_i
$$
- We use $\overline{\text{height}}$ as an estimate of the average population height.

---

## Confidence Interval for the Mean of a Continuous Variable

- Assume height is normally distributed with unknown mean but known variance $\sigma^2=16$, i.e.

$$
\text{height}_i \sim N(\mu, 16)
$$

---

## Confidence Interval for the Mean of a Continuous Variable

- The resulting 95% confidence interval for the average height in the population (assuming we know the population variance) is

```{r}
c(mean(height) - 
    qnorm(1 - (1 - 0.95) / 2) * sqrt(16 / n), 
  mean(height) + 
    qnorm(1 - (1 - 0.95) / 2) * sqrt(16 / n))
```

---

## Confidence Interval for the Mean of a Continuous Variable

- The math behind this code:

$$
\begin{aligned}
&\left[ \overline{\text{height}} - 1.96 *\sqrt{\frac{16}{n}},\quad  \overline{\text{height}} + 1.96 *\sqrt{\frac{16}{n}}\right] \\\\
= &\left[ 65.72 - 2*0.077, \quad 65.72 + 2*0.077\right] \\\\
= &\left[ 65.57, \quad 65.87 \right]
\end{aligned}
$$

---

## Confidence Interval for the Mean of a Continuous Variable

- Now assume we do not know either the population mean or the population variance

- You must estimate both from your sample

- The resulting 95% confidence interval is:

```{r}
c(mean(height) - 
    qt(1 - (1 - 0.95) / 2, n) * sqrt(var(height) / n),
  mean(height) + 
    qt(1 - (1 - 0.95) / 2, n) * sqrt(var(height) / n))
```

---

## Confidence Interval for the Mean of a Continuous Variable

- The math behind this code:

$$
\begin{aligned}
&\left[ \overline{\text{height}} - 1.96 * \sqrt{\frac{var(\text{height})}{n}},\quad  \overline{\text{height}} + 1.96 * \sqrt{\frac{var(\text{height})}{n}}\right] \\\\
= &\left[ 65.7 - 1.96*0.074, \quad 65.7 + 1.96*0.074\right] \\\\
= &\left[ 65.58, \quad 65.87 \right]
\end{aligned}
$$

---


## Estimating Population Proportion

- Estimate the proportion of older Americans who are female using the Alzheimer's data.

```{r}
mean(female)
```

- The math behind this code:

$$
\begin{aligned}
\overline{\text{female}} = \frac{1}{n} \sum_{i=1}^n \text{female}_i
\end{aligned}
$$

- We use $\overline{\text{female}}$ as an estimate of the proportion of females in the population, $p_\text{female}$.

---

## Confidence Interval for Proportion

- We assume that the variable __female__ follows a $\text{bernoulli}(p_{\text{female}})$ distribution

- The 95% confidence interval for $p_\text{female}$ is:

```{r}
c(mean(female) - qnorm(0.975) * sqrt(var(female) / n),
  mean(female) + qnorm(0.975) * sqrt(var(female) / n))
```

---

## Confidence Interval for Proportion

- The math behind the code:

$$
\begin{aligned}
&\left[ \overline{\text{female}} - 1.96*\sqrt{\frac{var(\text{female})}{n}},\quad  \overline{\text{female}} + 1.96*\sqrt{\frac{var(\text{female})}{n}}\right] \\\\
= &\left[ 0.57 - 1.96*0.01, \quad 0.57 + 1.96*0.01\right] \\\\
= &\left[ 0.56, \quad 0.59 \right]
\end{aligned}
$$

---

## Hypothesis Testing

- Test the hypothesis that the proportion of females in the population is less than 57%

- If $\bar{x}$ is our observed proportion and $\bar{X}$ is the sample proportion in a theoretical sample of n=2700, we are trying to find:

$$
P(\bar{X} \geq \bar{x} | H_0)
$$


---

## Hypothesis Testing

- There is a function in R that will perform this test


---

## Hypothesis Testing

```{r}
prop.test(sum(female), n=2700, p=0.57, 
          alternative = "less")
```

---

## Hypothesis Testing

- We now want to test a hypothesis about the population mean of a continuous variable

- Let $H_0$: the population mean csfvol is not equal to 350

- We use a t-test because we don't know the variance

---

## Hypothesis Testing

- There is a function in R that will perform this test

```{r}
t.test(data$csfvol, mu=350, 
       alternative="two.sided", conf.level=0.95)
```