---
title: "Confidence Interval and Hypothesis Testing"
author: "Yang Meng"
output: 
  xaringan::moon_reader:
    css: ["slide-style.css", "https://cdnjs.cloudflare.com/ajax/libs/font-awesome/5.15.3/css/all.min.css"]
    lib_dir: libs
    seal: false
    nature:
      beforeInit: "cols_macro.js"
      ratio: 16:9
      highlightStyle: "pygments"
      highlightLines: true
      highlightLanguage: "r"

---

class: title-slide


```{r echo = FALSE, message = FALSE}
library(tidyverse)
options(scipen = 999)
```


<br>
<br>
.right-panel[ 

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$author`


]


---

## Schedule for today

- Population Mean
- Population Proportion
- Confidence Interval
- Hypothesis Testing

---

## Load the Packages and Data

```{r, message=F, warning=F}
library(BSDA)
library(tidyverse)
df <- read.csv("../data/Alzheimer_data.csv")
```

---

## Simulated Data

```{r}
set.seed(1)
norm_size <- 100
norm_random <- rnorm(n = norm_size, mean = 10, sd = 2)
summary(norm_random)
```

---

## Confidence Interval for Population Mean

Suppose we know the population standard deviation:

```{r}
z_test_result <-
  z.test(x = norm_random, sigma.x = 2)
z_test_result$conf.int
```

---

## Confidence Interval for Population Mean

In general, however, the population standard deviation is unknown:

```{r}
t_test_result <- t.test(x = norm_random)
t_test_result$conf.int
```

---

## Adjust the confidence level

Note that we can give a narrower interval at a lower confidence level:

```{r}
t_test_result <- 
  t.test(x = norm_random, conf.level = 0.9)
t_test_result$conf.int
```

---

## Adjust the confidence level

On the other hand, increasing the confidence level leads to a broader interval:

```{r}
t_test_result <- 
  t.test(x = norm_random, conf.level = 0.99)
t_test_result$conf.int
```


---

## Hypothesis Testing

```{r}
summary(norm_random)
```

$H_0:\mu=10$ vs. $H_A:\mu\neq10$
```{r}
t.test(x = norm_random, mu = 10)$p.value
```

---

## Hypothesis Testing

$H_0:\mu=9$ vs. $H_A:\mu\neq9$
```{r}
t.test(x = norm_random, mu = 9)$p.value
```

---

## Practice 1

Try to get a 90 percent confidence interval for the population mean of brain volume (naccicv).

```{r}
summary(df$naccicv)
```

---

## Practice 1

Try to get a 90 percent confidence interval for the population mean of brain volume (naccicv).

```{r}
t_test_naccicv <- t.test(df$naccicv, conf.level = 0.9)
t_test_naccicv$conf.int
```

---

## Practice 2

Test whether the population mean is 1370 and get the p-value.

```{r}
summary(df$naccicv)
```

---

## Practice 2

Test whether the population mean is 1370 and get the p-value.

```{r}
t_test_naccicv <- t.test(df$naccicv, mu = 1370)
t_test_naccicv$p.value
```

---

## Simulated Data

```{r}
set.seed(1)
bern_size <- 100
bern_random <- rbinom(bern_size, 1, 0.3)
head(bern_random, 10)
success_prop <- sum(bern_random) / bern_size
success_prop
```

---

## Confidence Interval for Population Proportion

```{r}
prop_test_result <-
  prop.test(x = success_counts, 
            n = bern_size)
prop_test_result$conf.int
```

---

## Hypothesis Testing for Population Proportion

$H_0:p=0.3$ vs. $H_A:p\neq0.3$

```{r}
prop_test_result <-
  prop.test(x = success_counts, 
            n = bern_size, 
            p = 0.3)
prop_test_result$p.value
```

---

## Real Data Practice

```{r}
success_counts_naccicv <- 
  df %>%
  filter(naccicv > 1300 & naccicv < 1600) %>%
  nrow()
bern_size_naccicv <- nrow(df)
success_prop_naccicv <- 
  success_counts_naccicv / bern_size_naccicv
success_prop_naccicv
```

---

# Practice 

- Test whether the population proportion is 2/3.

- Get the 95 percent confidence interval for the population proportion.

---

```{r}
prop_test_result_naccicv <-
  prop.test(x = success_counts_naccicv, 
            n = bern_size_naccicv, 
            p = 2/3, 
            conf.level = 0.95)
prop_test_result_naccicv$conf.int
prop_test_result_naccicv$p.value
```

---

## Summary

- "t.test()" for population mean
- "prop.test()" for population proportion
- Useful arguments: "mu", "p", "conf.level"
- ?t.test and ?prop.test for more information


