---
title: "Confidence Interval and Hypothesis Testing"
author: "Yang Meng"
output: 
  xaringan::moon_reader:
    css: ["slide-style.css", "https://cdnjs.cloudflare.com/ajax/libs/font-awesome/5.15.3/css/all.min.css"]
    lib_dir: libs
    seal: false
    nature:
      beforeInit: "cols_macro.js"
      ratio: 16:9
      highlightStyle: "pygments"
      highlightLines: true
      highlightLanguage: "r"

---

class: title-slide


```{r echo = FALSE, message = FALSE}
library(tidyverse)
library(ggplot2)
options(scipen = 999)
```


<br>
<br>
.right-panel[ 

# `r rmarkdown::metadata$title`
## `r rmarkdown::metadata$author`


]


---

## Schedule for today

- Confidence interval for the population mean
- One-sample t-test
- Two-sample t-test
- Correlation test
- $\chi^2$ test

---

## Load the Packages and Data

```{r, message=F, warning=F}
library(tidyverse)
df <- read.csv("../data/Alzheimer_data.csv")
```

---

## Simulated Data

```{r}
set.seed(0)
norm_size <- 100
norm_random <- rnorm(n = norm_size, mean = 10, sd = 2)
summary(norm_random)
```

---

## Confidence Interval

```{r}
t_test_result <- t.test(x = norm_random)
t_test_result$conf.int
```

---

## Adjust the confidence level

We can lower our confidence level, which leads to a narrower interval:
```{r}
t_test_result <- 
  t.test(x = norm_random, conf.level = 0.9)
t_test_result$conf.int
```


---
## Adjust the confidence level

To have a higher confidence level, we need a broader interval:
```{r}
t_test_result <- 
  t.test(x = norm_random, conf.level = 0.99)
t_test_result$conf.int
```


---

## Hypothesis Testing

```{r}
summary(norm_random)
```

$H_0:\mu=10$ vs. $H_A:\mu\neq10$
```{r}
t.test(x = norm_random, mu = 10)$p.value
```

---

## Hypothesis Testing (One-sided)

$H_0:\mu=10$ vs. $H_A:\mu>10$
```{r}
t.test(x = norm_random, mu = 10, 
       alternative = "greater")$p.value
```

$H_0:\mu=9$ vs. $H_A:\mu>9$

```{r}
t.test(x = norm_random, mu = 9, 
       alternative = "greater")$p.value
```

---

## Practice 1: Confidence Interval

Try to get a 90 percent confidence interval for the population mean of age among AD subjects.

```{r}
summary(df$age)
```

---

## Practice 1: Confidence Interval

Try to get a 90 percent confidence interval for the population mean of age among AD subjects.

```{r}
t_test_age <- t.test(df$age, conf.level = 0.9)
t_test_age$conf.int
```

---

## Practice 2: One Sample t-test

Test whether the population mean of age is 70 or greater than 70 and get the p-value.

$H_0:\mu=70$ vs. $H_A:\mu>70$

```{r}
summary(df$age)
```

---

## Practice 2: One Sample t-test

Test whether the population mean of age is 70 or greater than 70 and get the p-value.

$H_0:\mu=70$ vs. $H_A:\mu>70$

```{r}
t_test_age <- t.test(df$age, mu = 70, alternative = "greater")
round(t_test_age$p.value, 4)
```

---

## Comparing Two Samples

Is blood pressure associated with gender?

```{r, echo = F, fig.height = 6, fig.width = 6, fig.align = "center"}
ggplot(data = df, 
       aes(x = factor(female), y = bpsys)) + 
  geom_boxplot() + 
  labs(x = "female")
```


---

## Two Sample t-test

We can examine whether the average blood pressure is different between male and female? Note that the boxplots above show the medians not the means.

$H_0:\mu_M=\mu_F$ vs. $H_A:\mu_M \neq \mu_F$ 

```{r}
t.test(bpsys ~ female, data = df)$p.value
```

---

## Two Numerical Variables

Question: Are age and blood pressure correlated?


```{r, echo = F, fig.height = 6, fig.width = 6, fig.align = "center"}
ggplot(data = df, 
       aes(x = age, y = bpsys)) + 
  geom_point()
```

---

## Correlation Test

$H_0:$ They are NOT correlated. vs. $H_A:$ They are correlated.

```{r}
round(cor.test(df$age, df$bpsys)$p.value, 4)
```

$H_0:$ They are NOT correlated. vs. $H_A:$ They are positively correlated.

```{r}
round(cor.test(df$age, df$bpsys, 
               alternative = "greater")$p.value, 4)
```


---

## Two Categorical Variables

Question: Are gender and disease status associated with each other?

```{r}
contingency_table <- table(df$female, df$diagnosis)
contingency_table
```

---

## Pearson's $\chi^2$ Test of Independence

$H_0:$ They are independent vs. $H_A:$ They are NOT independent.

```{r}
chisq.test(contingency_table)
```



---

## Summary

- "t.test()" for one/two-sample t-test
- "cor.test()" for correlation test
- "chisq.test()" for $\chi^2$ test
- Useful arguments: "mu", "conf.level", "alternative"
- ?t.test, ?cor.test and ?chisq.test for more information


