---
title: "Logistic Regression in R"
format:   
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    number-sections: true
    embed-resources: true
editor: visual
---
 
```{r}
#| echo: false
knitr::opts_chunk$set(warning = FALSE)
```

## Logistics Regression Model

* Remember from the lecture that we are fitting a regression model with a binary outcome. 

* As such, the model is as follows:

$$\begin{eqnarray*}
\log \Big(\frac{\hat{p}}{1 - \hat{p}} \Big) & = & a + b_{1}x_1 + \ldots + b_{q}x_{q}
\end{eqnarray*}$$


* The left hand side of this model is the logarithm of the odds of success.

* Thereby, the probability of success of $\pi$ can be written as follows:

$$\begin{eqnarray*}
\hat{p} & = & \frac{\exp(a + b_{1}x_1 + \ldots + b_{q}x_{q})}{1 + \exp(a + b_{1}x_1 + \ldots + b_{q}x_{q})}
\end{eqnarray*}$$

* The above means once we estimate the coefficients of the model, we can estimate the probability of success of the outcome of interest.


## Import data

* Let's revisit Alzheimer's data set, and consider the task of building a logistic regression model with diagnosis as its response variable and variables age, education, hippocampus volume, and female as its predictors.


```{r message=FALSE}
#| code-fold: TRUE
#| output-fold: TRUE

library(tidyverse) 

alzheimer_data <- read_csv("https://raw.githubusercontent.com/COSMOS-DataScience/slides/main/data/alzheimer_data.csv")

```
## Data pre-processing 

### Clean up the columns

* Running a logistic regression model in R is pretty straightforward. Before we do that, let's use `select()` function to select the interested variable (diagnosis, age, education, lhippo, rhippo, and female).


```{r}
#| code-fold: TRUE
#| output-fold: TRUE
alzheimer_data <- alzheimer_data %>% 
  select(diagnosis, age, educ, female, lhippo,rhippo)

glimpse(alzheimer_data)
```


### Data wrangling

* Let's begin by transforming the response to a new feature with two categories: no symptoms (0) versus mild or strong symptoms (1). 

* It is noticeable that female is a binary variable. As such, we should make sure R recognizes that feature as a factor variable.

* Create a new variable `hippo` as the sum of left hippocampus (lhippo) and right hippocampus (rhippo)

```{r}
#| code-fold: TRUE
#| output-fold: TRUE

alzheimer_data <- alzheimer_data %>% 
  mutate(diag = ifelse(diagnosis %in% c(1, 2), "1", "0"),
         diag = as.factor(diag),
         female = as.factor(female),
         hippo = lhippo + rhippo)
```

### Data visualization

* What is the relationship between diagnosis v.s. hippocampus?

* What is the relationship between diagnosis v.s. education years?

```{r}
#| code-fold: TRUE  

alzheimer_data %>% 
  ggplot(aes(x = hippo, y = diag, color = diag))+
  geom_boxplot()+
  theme_classic() +
  labs(x = 'Hippocampus volume (cc)', y = 'Diagnosis group')+
  theme(aspect.ratio = 0.5)

alzheimer_data %>% 
  mutate(female = fct_recode(female, Male="0",Female="1")) %>% 
  ggplot(aes(x = female, fill = diag))+
  geom_bar(position = 'fill') +
  theme_classic() +
  labs(x = 'Female', y = 'Diagnosis group')+
  theme(aspect.ratio = 0.5)


alzheimer_data %>% 
  ggplot(aes(x = educ, y = diag, color = diag))+
  geom_boxplot()+
  theme_classic() +
  labs(x = 'Education in years', y = 'Diagnosis group')+
  theme(aspect.ratio = 0.5)
```



## Model fitting

We will use `glm()` function to fit logistics regression.

```{r warning=FALSE}
#| code-fold: TRUE
#| output-fold: TRUE

logistic_model <- glm(
  diag ~ educ + age + female + hippo, 
  family = "binomial", 
  data = alzheimer_data)

summary(logistic_model)
```
 

## Model interpretation:

```{r echo=F}
betas = logistic_model$coefficients %>% round(digits = 2)
```

If we plug in the estimated values from the model above, we have

$$\begin{eqnarray*}
\log \Big(\frac{\hat{p}}{1- \hat{p}} \Big) & = & `r betas[1]` + `r betas[2]`\,x_{\text{educ}} + `r betas[3]`\,x_{\text{age}} + `r betas[4]`\,x_{\text{naccicv}} + `r betas[5]`\,x_{\text{female}}
\end{eqnarray*}$$

Let's say we want to estimate the probability of having symptoms for Alzheimer's disease for a person with specific characteristics. 

* What is the probability of having symptoms for Alzheimer's disease for a 40 years old women, with 10 education year, and 6 cc of hippocampus volume?

```{r}
newx_df <- data.frame(
  age    = 40,
  female = factor(0, levels = levels(alzheimer_data$female)),
  educ   = 10,
  hippo  = 6
) 
pred_prob = predict(logistic_model, newdata = newx_df, type = "response") %>%
  round(digits = 2)
```

$$\begin{eqnarray*}
\hat{P}(\text{diag = 1}) & = & \frac{\exp(`r betas[1]` + `r betas[2]`\,\times 10 + `r betas[3]`\,\times 40 + `r betas[5]` \times 1)}{1 + \exp(`r betas[1]` + `r betas[2]`\,\times 10 + `r betas[3]`\,\times 40 + `r betas[5]` \times 1)}
\end{eqnarray*} = `r pred_prob`$$

## Format the output to a publication-ready table

```{r warning=FALSE}
#| code-fold: TRUE
#| output-fold: TRUE
library(gtsummary) # for tbl_regression() function

logistic_model %>%
  tbl_regression(
    estimate_fun = function(x){style_number(x, digits = 3)}, 
    exponentiate = TRUE)
```
 

 
## Model Evaluation:

To split the data into training and validation sets using the rsample package in R, you can use the initial_split() function. Here's an example of how you can split the data:

```{r message=FALSE, warning=FALSE}
#| code-fold: TRUE
#| output-fold: TRUE
library(rsample)

set.seed(0)
data_split <- initial_split(alzheimer_data, prop = 0.7) 

train_data <- training(data_split)
test_data <- testing(data_split)
```

#### Fit the model with train data

As we saw, next step after splitting the data into train and test would be training the model using training data:

```{r warning=FALSE}
#| code-fold: TRUE
#| output-fold: TRUE
logistic_model2 <- glm(diag ~ educ + age + hippo + female, family=binomial, data=train_data)
summary(logistic_model2)
```


```{r warning=FALSE}
#| code-fold: TRUE
#| output-fold: TRUE
logistic_model2 %>%
  tbl_regression(estimate_fun = function(x) style_number(x, digits = 3), exponentiate = TRUE)
```


#### Validate on test data
Followed, by testing it via the validation set. This means to calculate the probability of success for each subject in the test set:


```{r warning=FALSE}
#| code-fold: TRUE
#| output-fold: TRUE
pred_prob <- logistic_model2 %>% 
  predict(test_data,type="response")
```


* We are now ready to calculate the accuracy of our trained model. To accomplish this, we translate all probabilities of success above to 0.5 to a 1 and otherwise to a 0, followed by tracking the number of correct predictions (1's correctly predicted as 1's and 0's correctly predicted as 0's).

```{r}
#| code-fold: TRUE
#| output-fold: TRUE

predicted.classes <- ifelse(pred_prob > 0.5, "1", "0")
acc = mean(predicted.classes == test_data$diag) %>% 
  round(2)

print(paste("This model yields a", acc, "accuracy rate!"))
```

