---
title: "Lab 5 Inference in R"
author: "<br>Data Science in Biomedical Sciences<br><br>UCI COSMOS 2026<br>Noah Benjamin<br>(adapted material from and Mine Dogucu)"
footer: ""
format:
  revealjs:
    chalkboard: true
    smaller: true
    slide-number: c/t
    code-line-numbers: false
    linestretch: 1.5
    html-math-method:
      method: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
---

```{r echo = FALSE, message = FALSE}
library(tidyverse)
library(ggplot2)
options(scipen = 999)
```

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.height = 9
)
ggplot2::theme_set(ggplot2::theme_light(base_size = 20))
```


## Schedule for today

- Confidence interval for the population mean
- One-sample t-test
- Test for Proportion
- Two-sample t-test
- Correlation test
- $\chi^2$ test


## Load the Packages and Data

```{r, message=F, warning=F}
library(tidyverse)
df <- read.csv("../../data/alzheimer_data.csv") 
# you probably have this and can name it whatever you like! 
# df stands for data.frame and its short, usually you want descriptive variable names
```


## Simulated Data

```{r}
set.seed(0) # We set a seed which makes the randomization work the same
norm_size <- 100
norm_random <- rnorm(n = norm_size, mean = 10, sd = 2) # simulate 100 draws from a N(10, 2) distribution
summary(norm_random) # let's see some summary statistics about our generated data
```


## Quick Confidence Interval Review

```{r}
# t-test that the mean of the data = 0
# we can add mu = 10 to change the mean we want to test for
t_test_result <- t.test(x = norm_random) 

# in this case we don't really care about the hypothesis 
# we just want a confidence interval for the mean
t_test_result$conf.int 
```


## Adjust the confidence level

We can lower our confidence level, which leads to a narrower interval:
```{r}
# we adjust the confidence level with the conf.level argument
# if we lower the confidence level it leads to a narrower interval
t_test_result <- t.test(x = norm_random, conf.level = 0.9) 
t_test_result$conf.int
```


## Adjust the confidence level

To have a higher confidence level, we need a broader interval:
```{r}
# if we increase the confidence level we widen the interval
t_test_result <- 
  t.test(x = norm_random, conf.level = 0.99)
t_test_result$conf.int
```



## Hypothesis Testing

We mentioned hypothesis testing earlier, let's try and perform a hypothesis test in R.

```{r}
summary(norm_random)
```

$H_0:\mu=10$ vs. $H_A:\mu\neq10$
```{r}
t.test(x = norm_random, mu = 10)$p.value
```


## Hypothesis Testing (One-sided)

$H_0:\mu=10$ vs. $H_A:\mu>10$
```{r}
t.test(x = norm_random, mu = 10, 
       alternative = "greater")$p.value
```

$H_0:\mu=9$ vs. $H_A:\mu>9$

```{r}
t.test(x = norm_random, mu = 9, 
       alternative = "greater")$p.value
```


## Practice 1: Confidence Interval

Try to get a 90 percent confidence interval for the population mean of age among AD subjects.

```{r}
summary(df$age)
```



## Practice 1: Confidence Interval

Try to get a 90 percent confidence interval for the population mean of age among AD subjects.

```{r}
t_test_age <- t.test(df$age, conf.level = 0.9)
t_test_age$conf.int
```


## Practice 2: One Sample t-test

Test whether the population mean of age is 70 or greater than 70 and get the p-value.

$H_0:\mu=70$ vs. $H_A:\mu>70$

```{r}
summary(df$age)
```


## Practice 2: One Sample t-test

Test whether the population mean of age is 70 or greater than 70 and get the p-value.

$H_0:\mu=70$ vs. $H_A:\mu>70$

```{r}
t_test_age <- t.test(df$age, mu = 70, alternative = "greater")
round(t_test_age$p.value, 4)
```


## Confidence Interval for Population Proportion

Simulated data

```{r}
set.seed(0)
bern_size <- 100
bern_random <- rbinom(bern_size, 1, 0.3)
head(bern_random, 10)
success_counts <- sum(bern_random)
success_counts
```


## Confidence Interval for Population Proportion

```{r}
prop_test_result <-
  prop.test(x = success_counts, 
            n = bern_size)
prop_test_result$conf.int
```


## Hypothesis Testing for Population Proportion

$H_0:p=0.3$ vs. $H_A:p\neq0.3$

```{r}
prop_test_result <-
  prop.test(x = success_counts, 
            n = bern_size, 
            p = 0.3)
prop_test_result$p.value
```


## Real Data Practice (brain volume: naccicv)

```{r}
success_counts_naccicv <- 
  df %>%
  filter(naccicv > 1300 & naccicv < 1600) %>%
  nrow()
bern_size_naccicv <- nrow(df)
success_counts_naccicv
bern_size_naccicv
```


# Practice 

- Test whether the population proportion is 2/3.

- Get a 95 percent confidence interval for the population proportion.

. . . 

```{r}
prop_test_result_naccicv <-
  prop.test(x = success_counts_naccicv, 
            n = bern_size_naccicv, 
            p = 2/3, 
            conf.level = 0.95)
prop_test_result_naccicv$conf.int
prop_test_result_naccicv$p.value
```


## Comparing Two Samples

Is blood pressure associated with gender?

```{r, echo = F, fig.height = 6, fig.width = 6, fig.align = "center"}
ggplot(data = df, 
       aes(x = factor(female), y = bpsys)) + 
  geom_boxplot() + 
  labs(x = "female")
```


## Two Sample t-test

We can examine whether the average blood pressure is different between male and female? Note that the boxplots above show the medians not the means.

$H_0:\mu_M=\mu_F$ vs. $H_A:\mu_M \neq \mu_F$ 

```{r}
t.test(bpsys ~ female, data = df)$p.value
```


## Two Numerical Variables

Question: Are age and blood pressure correlated?


```{r, echo = F, fig.height = 6, fig.width = 6, fig.align = "center"}
ggplot(data = df, 
       aes(x = age, y = bpsys)) + 
  geom_point()
```


## Correlation Test

$H_0:$ They are NOT correlated. vs. $H_A:$ They are correlated.

```{r}
# here, we are testing for a r = 0
cor.test(df$age, df$bpsys)$p.value
```

$H_0:$ They are NOT correlated. vs. $H_A:$ They are positively correlated.

```{r}
# here, we do the same but we only test for a positive association using a one-sided test
cor.test(df$age, df$bpsys, alternative = "greater")$p.value
```


## Two Categorical Variables

Question: Are gender and disease status associated with each other?

```{r}
contingency_table <- table(df$female, df$diagnosis)
contingency_table
```


## Pearson's $\chi^2$ Test of Independence

$H_0:$ They are independent vs. $H_A:$ They are NOT independent.

```{r}
# this is similar to cor.test, but for categorical variables instead of numeric ones
chisq.test(contingency_table)
```


## Summary

- "t.test()" for one/two-sample t-test
- "prop.test()" for proportion
- "cor.test()" for correlation test
- "chisq.test()" for $\chi^2$ test
- Useful arguments: "mu", "conf.level", "alternative"
- ?t.test, ?prop.test, ?cor.test and ?chisq.test for more information



# More hypothesis testing

Let's do a little data transformation first

```{r}
alzheimer_data <- read.csv('../../data/alzheimer_data.csv') %>% 
  select(id, diagnosis, age, educ, female, height, weight, lhippo, rhippo) %>% 
  mutate(diagnosis = as.factor(diagnosis), 
         female = as.factor(female),
         hippo=lhippo + rhippo)


alzheimer_healthy <- read.csv('../../data/alzheimer_data.csv') %>% 
  select(id, diagnosis, age, educ, female, height, weight, lhippo, rhippo) %>% 
  mutate(diagnosis = as.factor(diagnosis),
         female = as.factor(female), 
         hippo=lhippo+rhippo) %>% 
  filter(diagnosis==0)
```


## EXAMPLE: Two-Sample t-test

Ex. Do healthy men and women have the same mean left hippocampal volume?

-   Categorical (gender) and Numerical (Left Hippocampus) :

    -   Side by Side Boxplot

    -   Side by Side Violin

    -   2 Histograms

## Example: Visualization -- boxplot

```{r, fig.align='center'}
#| code-fold: true
#| echo: true

ggplot(data= alzheimer_healthy,
       mapping= aes(x= female, 
                    y=lhippo,
                    fill=female)) +
  geom_boxplot()+
  labs(
    title="Boxplot of lhippo by Gender",
    x="Gender",
    y="LHippo volume (cm^3)") +  
  theme_minimal()
```

## Example: Visualization -- violin

```{r, fig.align='center'}
#| code-fold: true
#| echo: true
ggplot(data= alzheimer_healthy,
       mapping= aes(x= female, 
                    y=lhippo,
                    fill=female)) +
  geom_violin()+
  labs(
    title="Violin of lhippo by Gender",
    x="Gender",
    y="LHippo volume (cm^3)") +  
  theme_minimal()
```

## Example: Visualization -- histogram

```{r, fig.align='center'}
#| code-fold: true
#| echo: true
ggplot(data= alzheimer_healthy,
       mapping= aes(x=lhippo,
                    fill=female)) +
  geom_histogram(bins=35)+
  labs(
    title="Histogram of Lhippo by Gender",
    x="LHippo volume (cm^3)",
    y="Count") 
```

## Example: use two-sample t-test

```{r, fig.align='center'}
#| code-fold: true
#| echo: true

t.test(alzheimer_healthy$lhippo[alzheimer_healthy$female==0], alzheimer_healthy$lhippo[alzheimer_healthy$female==1])
```


```{r, fig.align='center'}
#| echo: false
#alternative way to do the same test
t.test(alzheimer_healthy$lhippo ~ alzheimer_healthy$female)
```


## Example 2
- **Research Question**: Is the left hippocampal volume equal to the right hippocampal volume in humans?

- **Null Hypothesis**: Human’s left hippocampus volume is the same as the right hippocampus volume.

- **Alternative Hypothesis**: Human’s left hippocampus volume is not the same as the right hippocampus volume.

$$H_0: \mu_L = \mu_R \mbox{ vs } \mu_L \not= \mu_R$$

. . . 

- Method: **paired t-test**

## Example 2
- We essentially want to perform a two-sample t-test
- However, one fundamental assumption in the two-sample problem is that the two samples should be independent, such as the female group vs the male group. 
- Here we are looking at two features (left vs right hippocampus) of the same subjects. The correct test is a **paired t-test**, which is equivalent to perform a one-sample t-test using the differences


## Example 4
```{r}
#| code-fold: true
#| echo: true
t.test(alzheimer_data$rhippo, alzheimer_data$lhippo, paired = T)
```


```{r}
#| echo: false
#equivalently, 
t.test(alzheimer_data$rhippo- alzheimer_data$lhippo)
```


## Example 3

- **Research Question**:  Is having a large hippocampus (> 7 cm$^3$) associated with gender?

- This is a category variable vs category variable problem

. . . 

- Method: **chi-squared test**

```{r}
#| code-fold: true
#| echo: true
table(alzheimer_data$hippo>7,
      alzheimer_data$female)

chisq.test(table(alzheimer_data$rhippo>7,
      alzheimer_data$female))
```


## Example 4
- **Research Question**:  hippocampal volume correlated with age in healthy adults?

- **Null Hypothesis**: hippocampal volume and age are not correlated.
- **Alternative Hypothesis**: hippocampal volume and age are not correlated.

$$H_0: \rho=0 \mbox{ vs } H_1: \rho\not=0$$
. . . 

- Method: correlation test, more general, linear regression.

## Example 4: Visualization
```{r}
#| code-fold: true
#| echo: true
#| fig-align: center

#plot(alzheimer_healthy$age, alzheimer_healthy$hippo)
ggplot(data= alzheimer_healthy,
       mapping= aes(x= age, 
                    y= hippo)) +
  geom_point() +
  labs(
    title="Scatter plot of hippocampal volume vs age",
    x="Age (years)",
    y="Hippocampal volume (cm^3)") +  
  theme_minimal()
```

Discussion: is there a linear trend? are there regions of concerns? 


## Example 4
```{r}
#| code-fold: true
#| echo: true
cor.test(alzheimer_healthy$age, alzheimer_healthy$hippo)

#alternatively, we can use linear regression
summary(lm(alzheimer_healthy$hippo ~ alzheimer_healthy$age))
```


## Example 5
- **Research Question**: Does left hippocampal volume differ across diagnostic groups?

- **Null Hypothesis**: 

- **Alternative Hypothesis**:

$$H_0: \mu_0=\mu_1=\mu_2 \mbox{ vs } H_1: \mbox{ at least two means are different}$$

. . . 

- Method: one-way analysis of variance (ANOVA), more generally, linear regression


## Example 5
```{r, fig.align='center'}
#| code-fold: true
#| echo: true

ggplot(data= alzheimer_data,
       mapping= aes(x= diagnosis, 
                    y=lhippo,
                    fill=diagnosis)) +
  geom_boxplot()+
  labs(
    title="Boxplot of lhippo by Diagnosis",
    x="Diagnosis",
    y="LHippo volume (cm^3)") +  
  theme_minimal()
```

## 
```{r}
#| code-fold: true
#| echo: true
summary(aov(alzheimer_data$lhippo ~ alzheimer_data$diagnosis))

#equivalently
summary(aov(lhippo ~ diagnosis, data=alzheimer_data))
```

## Summary
::: {style="font-size: 70%;"}

- Ex 1. Is the hippocampal volume of healthy males greater than 6 cm$^3$? (one-sample t-test)

- Ex 2. Is the proportion of healthy adults with a right hippocampal volume > 3cm$^3$ equal to 50%? (one-sample proportion test or z-test)

- Ex 3. Do healthy men and women have the same mean left hippocampal volume? (two-sample t-test)

- Ex 4. Is the left hippocampal volume equal to the right hippocampal volume in humans?
(paired t-test)

- Ex 5. Is having a large hippocampus (> 7 cm$^3$) associated with gender?
(chi-squared test)

- Ex 6. Is hippocampal volume correlated with age in healthy adults?
(Pearson’s correlation or linear regression)

- Ex 7. Does left hippocampal volume differ across diagnostic groups?
(one-way ANOVA or linear regression)
:::

<!--
## Advanced: NONPARAMETRIC TESTS

::: {style="font-size: 75%;"}
- The tests we introduced might not work well for small sample sizes.

- There are corresponding parametric methods, which are more robust. 
 
1. one-sample t-test : signed rank test

2. two-sample t-test : Mann-Whitney U test, also known as Wilcoxon rank sum test

3. one-sample proportion test : binomial test

4. paired t-test : signed rank test

5. chi-squared test : Fisher’s exact test

6. Pearson’s correlation : Spearman’s correlation

7. one-way ANOVA : Kruskal-Wallis H test
:::

-->

