---
title: "Lab 1 intro to R"
author: "<br>Data Science in Biomedical Sciences<br><br>UCI COSMOS 2026<br>Noah Benjamin"
footer: ""
format:
  revealjs:
    chalkboard: true
    smaller: true
    slide-number: c/t
    code-line-numbers: false
    linestretch: 1.5
    html-math-method:
      method: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.height = 9
)
ggplot2::theme_set(ggplot2::theme_light(base_size = 20))
```

## Slides
- Slides can be found on the website (https://cosmos-datascience.netlify.app). 

## First things first:
- Do you have a computer?
- Do you have R on your computer?
- Do you have RStudio on your computer?


## Let's talk about RStudio

Why is it a nice environment for R coding??

. . .

- It is an IDE(Integrated Development Environment) that we use to write, manage, and execute R code.

. . .

- It is open-source and free!! (and works well too)

. . .

- It's not the only R-capable IDE, most of the popular ones have R extensions (like VSCode)


## Today's lab

We're going to get some practice with the basics of R.

. . . 

- The slides will walk through the concepts and serve as a reference. At the same time, we will each be writing the code on our computers to get some hands-on practice. I'll also be doing it!!

. . .

- Please stop and ask questions!! It doesn't have to just be if you're stuck. 


## First, create a Quarto document

This is like a notebook file where you can write both text and code. It's saved as a .qmd file.

. . . 

- Quarto can run both **R** and **Python**!!! We will only be focusing on R.

. . . 

- Also, it has built-in $\LaTeX$ (very cool)

. . . 

- $y0u\mathbb{R}\varepsilon$ $a\omega \varepsilon s0m\varepsilon$


## Lets start coding!

. . . 

Code is written in code chunks:

. . . 

```{r}
# Hi I am a code chunk
```

. . .

Code chunks are delineated by

```{r}
# Without the # mark:
# ```{r}
# ```
```

. . .

- To make a new code chunk, there is a handy keyboard shortcut 

. . . 

- (Ctrl + Alt + I)

- (Command + Option + I)


## Math in R

. . .

- We can perform traditional mathematical operations in R

. . . 

```{r}
3+3 # add 3+3=6
4-1 # subtract 4-1=3

3/3 # divide 3/3=1
4*1 # multiply 4*1=4

2^4 # we can use either ^ or ** for exponent
2**4 
```


## Math in R


- We can perform traditional mathematical operations in R

```{r}
15 %% 2 # 'modulo', the remainder when (in this case) 15 is divided by 2

sqrt(16)

log(100)

exp(4)
```


## Math in R

. . . 

Try and write your own to solve:

. . . 

$$\frac{e^4*|-ln(3)|}{-19*45}$$

. . . 

```{r}
(exp(4)*abs(-log(3)))/(-19*45)
```


## Creating a variable (or object)

. . . 

- The fancy term is the *Assignment operator*

. . . 

- Basically just means we assign a value to a variable with the name we specify for it

. . . 

```{r}
birth_year <- 2000 # this is my birth year, maybe fill in your own!
```

. . . 

- In this case we assigned the value 2000 to an object with the name `birth_year`

. . . 

- We can then display the value of this variable by just running the variable name (either in it's own chunk, or in the console)

```{r}
birth_year
```


## Comments

. . . 

- As you have already seen, we can write comments in our code chunks.

- We use the # symbol

. . . 

- They are just for our reference and do not do anything when we run the chunk

- Like writing little notes in the code

. . . 

```{r}
my_age <- 2026 - birth_year # we can also do math in the code!!
```


## R is case-sensitive

. . . 

- Let's say I want to display the variable I created on the previous slide

. . . 

```{r, eval=FALSE}
My_age
```

. . . 

- R doesn't like that because it's very picky

- you can only reference the variable the *exact* way it was defined

. . . 

```{r}
my_age
```


## Some other variable naming conventions/restrictions

. . . 

Allowed Characters:

- letters (`a-z`, `A-Z`), numbers (`0-9`), dots (`.`), and underscores (`_`)

. . . 

Forbidden Characters: 

- spaces and some special characters (`!`, `@`, `#`, `$`, `%`, `-`)

. . . 

Starting Character: 

- variables names must start with a letter or a period (`.`)

- don't start names with an underscore or a number

- can't name a variable with a number right after a period (ex. `.2var` is invalid)


## Variables in R

. . . 

Variables are categorized into different types. You might be familiar with some of these already:

. . . 

- `character`: takes string values (e.g. a person's name, address)

. . . 

- `integer`: integer (single precision)  

. . . 

- `double`: floating decimal (double precision) 

. . . 

- `numeric`: integer or double  

. . . 

- `factor`: categorical variables with different levels (ex. knitting skill: low, medium, high)

. . . 

- `logical`: TRUE (1 or T), FALSE (0, or F)  

. . . 



## Variables in R

If we have a variable we want to check the type of, we use the `typeof()` function.

. . . 

```{r}
typeof(1) # by default most numbers are stored as doubles
typeof(1L) # using the L suffix forces R to store it as an integer
typeof(1.5)
typeof(birth_year) # most of the time you will use typeof() on a variable like this
typeof(as.integer(1))
```


## A little more on `character` variables

. . .

- Anything we define in quotes is a string (even if it's all numbers like "1234").

. . .

- It's how we differentiate variables names and text

. . . 

```{r}
"my_age" # is just the characters itself

my_age # is the variable with the value we previously assigned to it
```


## Vectors (or lists)

. . . 

- Often, we want to store ordered lists of values, which we call vectors
- We call each component of the vector an *element*

. . . 

- This is done using the `c()` function, which "concatenates" the values together into one vector

. . . 

- Let's create some vectors about fruit

. . . 

```{r}
fruits <- c("Apple", "Banana", "Orange")

colors <- c("red", "yellow", "orange")

seeds <- c(T, F, T)
```


## Vectors (or lists)

. . .

- We can access specific elements in these vectors through *indexing*

. . . 

- We index through the elements of a vector using the brackets `[]`

. . . 

- So, if we want to access the first element of the `fruits` vector we write: 


```{r}
fruits[1]
```

. . . 

- R is a 1-indexed language, meaning that when we index through a list we start at 1

- If you are familiar with Python, that is a 0-indexed language



## data.frame

. . .

- One of the most common ways we will store data is in a data frame

. . . 

- It's like a spreadsheet, that stores data in a grid format

- In this grid each column (or row) is its own vector

. . . 

- We can use our vectors about fruit to create a `data.frame` that stores all the info about fruit

. . . 

```{r}
fruit_data <- data.frame(fruit = fruits, color = colors, seeds = seeds)
fruit_data
```

## data.frame

- We can index the `data.frame` grid with brackets as well, indexing through the rows and columns

```{r, eval = FALSE}
fruit_data[row, column]
```

. . . 

```{r}
fruit_data[2, 1] # this will give us the second row, first column entry of fruit_data
```

. . . 

- We can leave one of the row or column entries blank if we want the entire row or column

```{r}
fruit_data[,1] # will give us the entire first column

fruit_data[3,] # will give us the entire third row
```



## Functions

. . . 

- Let's say we have a list of numbers, and we want to find the average of them

```{r}
numbers <- c(-1, 15, 100, 111, 110)
```

. . . 

- We could this this mathematically, by adding up the values and dividing by how many there are

```{r}
(-1 + 15 + 100 + 111 + 110)/5
```

## Functions

- R has a lot of handy ways of making this quicker for us, however

. . . 

- The `sum()` function will take the sum of the elements of the vector we pass into it
- The `length()` function will tell us how many elements are in the vector we give it

```{r}
total <- sum(numbers)
how_many <- length(numbers)
```

. . . 

```{r}
total/how_many # so now we can find the mean this way
```


## Functions

. . . 

- Conveniently, R has a `mean()` function, which will do all of these things for us

. . . 

```{r}
mean(numbers)
```


## Functions

Generally, we call functions in the following way:

. . . 

```{r, eval=FALSE}
do(something)
```

`do()` is a function;   
`something` is the argument of the function.


## Functions 

Functions are not limited to only one argument:

```{r, eval=FALSE}
do(something, colorful)
```

`do()` is a function;   
`something` is the first argument of the function;   
`colorful` is the second argument of the function.

. . .

- Some arguments might be optional, and start with default values


## Functions

To learn about what a function does, we can use a question mark before the function:

. . . 

```{r}
?mean
```

. . . 

- This will tell us what arguments a function expects, and default values for optional ones


## Functions

- Many functions exist either in base R or in packages that save us from having to write our own code

. . . 

- If you are looking for something specific, Google is your friend!!


## Conditional statements

. . . 

```{r, eval=FALSE}
if(condition){
   Do Something
} else {
   Do Something Else
}
```

. . . 

- We can add more conditions that are checked sequentially


## Conditional statements

```{r, eval=FALSE}
if(condition1){
  Do something
} else if (condition2) {
  Do a different thing
} else {
  Do this different thing
}
```

. . . 

- Only one of the blocks will be run depending on which condition is met

## Writing a condition

. . . 

- `==` to check if something is equal (`x == 2`)

. . . 

- `!=` to check if it's not equal to

. . . 

Traditional inequalities also work:
- `<`, `>`, `<=`, `>=`


## Writing a condition

- We can also combine multiple conditions, to only execute if both conditions are met, or if one of two is met

. . . 

- `&&` check for both conditions to be true

- ex: `if(x == 4 && y < 2)`, only returns TRUE if both conditions are met 

. . . 

- `||` checks for at least one of the conditions to be true

- ex: `if(x == 4 || y < 2)`, returns TRUE if either of the two conditions are met 


## Conditional example 1

```{r}
height <- 63
if(height >= 63){ # here we check if height is large enough to take the ride
  print("Fantastic! Enjoy the rollercoaster")
} else {
  print("OOPS! Sorry, you'll have to wait until next year")
}
```


## Conditional example 2

```{r}
money <- 10
greed <- 1
if(money-1000 < 0 || greed > 50){ # here we check if we don't have enough money, or we are too greedy to lend
  print("Sorry I don't have enough money to loan you")
}else{
  print("Sure, here's 1000$.")
}
```


Try and play around with the amount assigned to 'money' and 'greed'.


## Make your own

Now, Try and make your own if statement!

```{r, eval=FALSE}
if(){
  print("")
}else{
  print("")
}
```


## `while` loops and `for` loops

. . . 

- `while` loops and `for` loops, as the names suggest repeat (or 'loop') a segment of code over and over

. . . 

- a `while` loop runs code *while* a certain condition is met

. . . 

- a `for` loop runs code *for* a pre-specified number of iterations



## `while` loop syntax

```{r, eval = FALSE}
while(condition) {
  Do this over and over 
}
```

. . . 

- Be careful because its easy to write a `while` loop that runs for ever (and that is bad)


## `while` loop example


```{r, eval = FALSE}
x <- 0
while(x < 10) {
  x <- x + 1
}
```

. . . 

-  We can put any code we'd like in the loop


## `while` loop example

```{r}
x <- 0
while(x < 10) {
  x <- x + 1
  if(x %% 2 == 0) { # check if x is even
    print(x)
  }
}
```


## `while` loop example

We can also forcibly break out of a loop when a certain condition is met, even if the loop has no actual end

. . . 

```{r}
x <- 1
while(TRUE) {
  if (x == 5) {
    break
  }
  print(x)
  x <- x + 1
}
```


## `for` loop

. . . 

- remember that `for` loops repeat for a specified number of iterations

. . . 

- in the `while` loop example we just saw, we manually increased `x` by 1 every iteration of the loop

. . . 


- `for` loops are special because they do this automatically


## `for` loop syntax

```{r, eval = FALSE}
for(element in vector){
  Do this as many times as there are elements in vector
}
```

. . . 

- how this looks in practice:

## `for` loop example

```{r}
numbers <- c(1, 2, 3, 4, 5)
for(num in numbers) {
  print(num)
}
```
. . . 

- one of the nice things about this is that the loop will iterate over any list, not just numbers

## `for` loop example

```{r}
for(fruit in fruits){ # using the fruits vector we defined earlier
  print(fruit)
}
```


## Creating Functions

Remember the `length()`, `sum()`, and `mean()` functions we saw earlier?

. . . 

What if we want to make our own functions in R?

. . . 

Writing your own functions is useful when you find you are rewriting the same code over and over again. 

. . . 

- We learned that functions take in *arguments* that are the user-inputted values when they call the function/

. . . 

- The other indispensable ingredient in a function is a `return()` statement

- When a function hits `return(some_value)`, it immediately ends and outputs the return value


## Function syntax

Functions are made using the following syntax, where it's important to prespecify the input arguments, and how you want them processed.

```{r}
my_function <- function(input){
  if(input == TRUE){
    return("Fantastic!")
  }
  else{
    return("OOPS!")
  }
}
```

. . . 

- What does this function take in, what are the arguments?

- What does this function do with those arguments?

What kind of inputs would return either result?


## Running the function

When we run the `my_function()` code, we just created the function

. . . 

- If we actually want to run it, we have to call it like we called the previous functions

. . . 

```{r}
ate_lunch <- TRUE
my_function(ate_lunch)
```

. . . 

- We must first define the input variable, and then pass it into the function as an argument


## Let's write our own `mean()` function

```{r}
my_mean <- function(numbers) {
  total <- sum(numbers)
  how_many <- length(numbers)
  return(total/numbers)
}
```


## R packages

Sometimes, other people write functions that are useful to us. This is where packages come in.

. . . 

A package is essentially a library of functions written by someone and compiled for a specific purpose

. . .

We use packages by first installing them, and then loading them into our R environment


## R packages

Let's ry running the following code:

```{r}
beep()
```

. . . 

Why are we seeing an error?

. . . 

Answer: We don't have this package downloaded.

Think of packages as items, when we have an error it means we have yet to get/purchase an item.


## Installing 'Getting' a Package

You can copy or type install.packages("beepr") in the console, or in the a code chunk.

NOTE: we only ever need to install a package once. Once downloaded, we can 'call' it and use it whenever we want.

```{r}
install.packages("beepr")
```

. . . 

We can also use the 'packages' search in the bottom-right pane


## Using Packages and Functions from Packages 

We want to use the beep() function from the beepr package. So we first need to call/open the package.

. . . 

How do we get it?

We use the function `library()` to recall and load in the package into our R session.

```{r}
library("beepr")
```

. . . 

Now try and run `beep()`

```{r}
beep()
```

. . . 

YAY!

## We can still use `?` to get help about a function

What if we want a different sound? Where do we go again if we don't know what to do, or maybe want to learn more about a particular function?

Use help search bar, or '?'

```{r}
?beep
```

. . . 

Good packages have documentation written about what the functions do

## We can still use `?` to get help about a function

What kinds of sounds are there from the description?

- 1. ping
- 2...
- 3...

- Try out different sound types

```{r}
beep(sound="wilhelm")
```


## More Fun

Now, use the say() function from the library cowsay and see what it does

. . . 

```{r}
install.packages("cowsay")
library(cowsay)
```
. . . 

Need Help?
```{r}
?say()
```

- It will Print out default "Hello world", and generate an image of a cat
- we can change what it says by adjusting the 'what' argument
- we can change the animal by adjusting the 'by' argument

. . . 

What are some of the animals it can generate?

Are there other things you can change?

. . . 

Play around!

. . .

```{r}
say(what="Have Fun!", by="cat", what_color="green")
```


